---
title: "summarize capture data"
author: "Annemarie van der Marel"
date: "2023-10-03"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

sessionInfo()
```

# import data

```{r import}

trapdata <- read.csv("./data/degu_capture_09-20.csv") %>% dplyr::select(-X)
    #,col_types = cols(date = col_date(format = "%Y-%m-%d"))) 

str(trapdata)


```

# select first capture

```{r}

unique(trapdata$capture)
trap <- trapdata %>%
  filter(capture==1)

unique(trap$capture)
```



# seasons


```{r julian dates,  eval=FALSE, include=FALSE } 
dates <- read_csv("./data/dates.csv", col_types = cols(date = col_date(format = "%Y-%m-%d"))) 

 trap_dates <- trap %>% select(julian_date, date) %>% distinct() 

 dates <- dates %>% left_join(trap_dates)  


``` 

```{r number of days, eval=FALSE, include=FALSE} 

 maxdays <- dates %>%  
  pivot_wider(id_cols=period, 
              names_from=time,  
              values_from = julian_date  
             ) %>% 
  mutate(length=end-start) 

 maxdays 
```






# summarize capture dates


```{r field season}

fieldseason <- trap %>%
  group_by(season, year) %>%
  summarize(start = min(date), end= max(date)) %>%
  mutate(status="trapping")  


```


## fall distribution

```{r fall trapping distribution}

fall <- trap %>%
  filter( season=="fall") %>% 
  group_by(year) %>%
  summarize(start=min(date), end=max(date),
            start.jd=min(julian_date), end.jd=max(julian_date))
fall

trapdist_fall <- trap %>%
  filter( season=="fall") %>%  
  ggplot(aes(julian_date)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right") +
  scale_x_continuous(limits = c(124, 203), n.breaks = 8 ) +
  geom_vline(xintercept = c(155, 174), color="orange")
trapdist_fall 

ndays_fallwindow<- 174-155

```

fall dates trapping:
2013-06-03 = julian 155
2009-06-23 = julian 174

max of 19 day period , constraint by trapping in 2009 and 2013 but we can make the window larger as 
this period data are too sparse. 

## spring distribution
```{r spring trapping distribution}
spring <- trap %>%
  filter(season=="spring") %>% 
  group_by(year) %>%
  summarize(start=min(date), end=max(date),
            start.jd=min(julian_date), end.jd=max(julian_date))
spring


trapdist_spring <- trap %>%
  filter(season=="spring") %>%
  ggplot( aes(julian_date)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right") +
  scale_x_continuous(limits = c(230, 320), n.breaks = 9 ) +      # fall trapping
  geom_vline(xintercept = c(244, 300), color="lightgreen") 
trapdist_spring

ndays_springwindow<- 300-244

```


spring dates trapping:
2009-09-01 = julian 244
2013-10-26 = julian 300

max of 59 days


## breeding season
Look at dates of mating, gestation, lactation, first day of juvenile emergence


```{r}
levels(as.factor(trap$breeding_status_2))
#levels(as.factor(trap$notes))
```
We do not have much data on mating, just one copulatory plug observation. 


```{r gestation}

levels(as.factor(trap$breeding_status_2))

check_gestation <- trap %>%
  group_by(breeding_status_2) %>%
  tally()
write.csv(check_gestation, "breeding_status.csv") 


gestation <- c("Preñada", "Prenada?","prenada?", "P", "preñada", "preñada?" , "preñada???" )

check_pregnancy <- trap %>%
  filter(breeding_status_2 %in% gestation)

pregnancy <- trap %>%
  filter(breeding_status_2 %in% gestation) %>%
  group_by(season, year) %>%
  summarize(start = min(date), end= max(date),
            start.jd=min(julian_date), end.jd=max(julian_date)) %>%
  mutate(status="gestation") 



```
Gestation start, at least observed, was in fall earliest on 2018-07-02 (julian 185) and latest 	
2017-07-11 (julian 194). 

first in spring was 2018-08-10 (224) and the last one in spring was 2009-10-25 (julian 298). 


```{r lactation}
levels(as.factor(trap$breeding_status_2))

milk <- c("lactante reciente?", "lactante, X", "lactante", "lactante, no preñada", 
          "lactante, preñada?", "lactante?, X", "lactante reciente", "lactante, no preñada?",
          "lactante, secresión sang.")

lactation <- trap %>%
  filter(breeding_status_2 %in% milk) %>%
  group_by(season, year) %>%
  summarize(start = min(date), end= max(date),
            start.jd=min(julian_date), end.jd=max(julian_date)) %>%
  mutate(status="lactation") 


```

Lactation start, at least observed, was latest on 2011-09-14 (Julian 257) and stopped earliest on 2010-10-19 (julian 292. 


```{r juvenile emergence}

levels(as.factor(trap$status))
levels(as.factor(trap$age))

# pupemerge <- trap %>%
#   filter(notes=="pup") %>%
#   arrange(julian_date) %>%
#   group_by(year) %>%
#   slice(1, n()) %>%
#   ungroup

pupemerge <- trap %>%
  filter(status=="new", grepl('pup', notes)) %>%
  group_by(season, year) %>%
  summarize(start = min(date), end= max(date),
            start.jd=min(julian_date), end.jd=max(julian_date)) 

check <- trap %>%
  filter(year==2017)
pupemerge.age <- trap %>%
  filter(status=="new", age %in% c("P", "Pup")) %>%
  group_by(season, year) %>%
  summarize(start = min(date), end= max(date),
            start.jd=min(julian_date), end.jd=max(julian_date)) 

pups <- bind_rows(pupemerge, 
                  pupemerge.age) %>%
  mutate(status="juvenile_emergence") 
  
  
```

pup emergence start was latest on 2019-10-04 (julian 279) . 

```{r weaning}

postlactation <- c("post lactante, no preñada?", "postlactante, preñada", 
                   "post lactante, preñada?", "post lactante", "post lactante, no preñada?!",
                   "post lactante, X", "postlactante", "postlactante, preñada?", 
                   "post lactante, no preñada", "post lactante?", "postlactante, no preñada", 
                   "postlactante, X" )

weaning <- trap %>%
  filter(breeding_status_2 %in% postlactation) %>%
  group_by(season, year) %>%
  summarize(start = min(date), end= max(date),
            start.jd=min(julian_date), end.jd=max(julian_date)) %>%
  mutate(status="weaning") 

```
weaning start, at least observed, was latest on 2009-10-21 (294) and stopped earliest on 2013-10-26 (300). 


```{r combine dates}
 
fieldseason_dates <- bind_rows(fieldseason, 
                              pregnancy, 
                              lactation, 
                              pups, 
                              weaning)

write.csv(fieldseason_dates, "./data/fieldseason_dates.csv")

```

#total trap effort

```{r}
trap_effort_total <- trap %>%
  group_by(season, year) %>%
  summarize(total_days=length(unique(date)),
            total_ids=length(unique(ear_tag_id))) %>%
  pivot_longer(cols = c("total_days", "total_ids"), names_to = "trap", values_to = "total" )
trap_effort_total$trap[trap_effort_total$trap=="total_days"]<- "days"
trap_effort_total$trap[trap_effort_total$trap=="total_ids"]<- "ids"
# 2020, no degus captured
trap_effort_total$total[48]<- 0 #[trap_effort_total$year==2020,trap_effort_total$total==ids]
trap_effort_total$total[24]<- 0

trapeffort_wide <- trap_effort_total %>%
  pivot_wider(names_from = trap, 
              values_from = total)


write.csv(trapeffort_wide, "data/total trap effort.csv")

```



# Plot distribution of trapping


Distribution of trapping (histogram of trapping days) years 2009-2020

```{r plot}

trapdist <- trap %>%
  #filter(!year %in% c(2009, 2010)) %>%
  ggplot( aes(julian_date)) +
  geom_histogram(stat = "count") +
  labs(x="julian date") +
  #theme_classic()+
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1, strip.position = "right") +
  scale_x_continuous(limits = c(130, 320), n.breaks = 20 ) +
  geom_vline(xintercept = c(155, 174), color="orange") +      # fall trapping / select fall
  #geom_text(x=157.5, y = 120, label="fall", color="orange", size=3) +
  geom_vline(xintercept = c(244, 300), color="lightgreen") +  # spring trapping
  #geom_text(x=247, y = 120, label="spring", color="lightgreen", size=3) +
  geom_vline(xintercept = c(194, 248), color="pink") +        # gestation 
  #geom_text(x=197, y = 120, label="gest", color="pink", size=3) +
  geom_vline(xintercept = c(257, 292), color="lightblue") +   # lactation
  #geom_text(x=260, y = 120, label="lact", color="lightblue", size=3) +
  geom_vline(xintercept = c(279), color="yellow") +            # juvenile emergence
  geom_vline(xintercept = c(279, 298), color="red")            # select spring
  #geom_text(x=282, y = 120, label="pup", color="yellow", size=3) 
  #geom_vline(xintercept = c(294, 300), color="purple")        # weaning
trapdist

ggsave( "./data/plot_trapping distribution.png")

```




# 30-day periods

fall dates trapping:
2013-06-03 = julian 155
2009-06-23 = julian 174

choose 154 + 30 = 184

spring, perhaps from juvenile emergence onward
2019-10-04 = julian 279 + 30 = 309


## fall distribution

```{r fall select}

fall.select <- trap %>% 
  filter( season=="fall", between(julian_date, 154, 184)) %>%
  group_by(year) %>%
  summarize(start=min(date), end=max(date),
            start.jd=min(julian_date), end.jd=max(julian_date))
fall.select

select_fall <- trap %>%
  filter( season=="fall", between(julian_date, 154, 184)) %>%  
  ggplot(aes(julian_date)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right") +
  scale_x_continuous(limits = c(150, 190), n.breaks = 8 ) 
select_fall 


fall_distr <- trap %>%
  filter( season=="fall") %>%  
  ggplot(aes(julian_date)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right") 
   
fall_distr


```



## spring distribution
```{r spring select}

spring.select <- trap %>% 
  filter( season=="spring", between(julian_date, 279, 309)) %>%
  group_by(year) %>%
  summarize(start=min(date), end=max(date),
            start.jd=min(julian_date), end.jd=max(julian_date))
spring.select



spring_distr <- trap %>%
  filter(season=="spring") %>%
  ggplot( aes(julian_date)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right")  
spring_distr



select_spring <- trap %>%
  filter(season=="spring", between(julian_date, 279, 309)) %>%
  ggplot( aes(julian_date)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right") +
  scale_x_continuous(limits = c(275, 310), n.breaks = 9 ) 
select_spring



```

## Plot distribution of trapping


Distribution of trapping (histogram of trapping days) years 2009-2020 of the 2 30 day windows

```{r plot}

trapdist30d <- trap %>%
  #filter(!year %in% c(2009, 2010)) %>%
  ggplot( aes(julian_date)) +
  geom_histogram(stat = "count") +
  labs(x="julian date") +
  #theme_classic()+
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1, strip.position = "right") +
  scale_x_continuous(limits = c(130, 320), n.breaks = 20 ) +
  geom_vline(xintercept = c(154, 184), color="orange") +      # fall trapping / select fall
  #geom_text(x=157.5, y = 120, label="fall", color="orange", size=3) +
  geom_vline(xintercept = c(279, 309), color="lightgreen")   # spring trapping
  
trapdist30d

ggsave( "./data/plot_trapping distribution 30-days.png")

```

## visualize trap effort

fall<- filter( season=="fall", between(julian_date, 154, 184)
season=="spring", between(julian_date, 279, 309)


are the same individuals trapped outside and inside these windows?

```{r include trap window category}

trapwindow <- trap %>%
  group_by(year) %>%
  mutate(window = if_else(between(julian_date, 154, 184), "fall", 
                          if_else(between(julian_date, 279, 309), "spring", "exclude")))
```



By year and season, is the trapping effort (days of trapping) and number of trapped individuals  by all individuals similar? 

By window
```{r}

trap_effortXwindow <- trapwindow %>%
  group_by(season, year, window) %>%
  summarize(n_days=length(unique(date)),
            n_ids=length(unique(ear_tag_id))) %>%
  filter(window!="exclude") %>%
  arrange(year)
# 2020, no degus captured
trap_effortXwindow$n_ids[trap_effortXwindow$year==2020]<- 0


plotXwindow <- ggplot(trapwindow, aes(x=window)) +
  geom_histogram(stat = "count") +
  facet_wrap(~year, #*season
             nrow = 12, ncol = 1,
             strip.position = "right") 
plotXwindow 
```

By the two selected windows

```{r}
trap_effortXselectwindow <- trapwindow %>%
  group_by(season,  window, year) %>%
  #filter(window %in% c("mating", "lactation", "pups")) %>%
  summarize(n_days=length(unique(date)),
            n_ids=length(unique(ear_tag_id))) %>%
  pivot_longer(cols = c("n_days", "n_ids"), names_to = "trap", values_to = "count" ) 

trap_effortXselectwindow$trap[trap_effortXselectwindow$trap=="n_days"]<- "days"
trap_effortXselectwindow$trap[trap_effortXselectwindow$trap=="n_ids"]<- "ids"

# 2020, no degus captured
trap_effortXselectwindow$count[24]<- 0
trap_effortXselectwindow$count[48]<- 0
trap_effortXselectwindow$count[72]<- 0


# combine total trap effort and by window
trapeffort <- left_join(trap_effortXselectwindow, trap_effort_total, by=c("season", "year", "trap")) 

# plots
plotfall <- trapeffort %>%
  filter(window=="fall") %>% 
  ggplot( aes(x=year, y=count, fill=trap)) +
  geom_col(stat = "identity", position = "dodge") +
  labs(y="count", title = "fall") +
  theme_classic() +
  #scale_y_continuous(limits = c(0, 130) ) +
  scale_x_continuous(limits = c(2008, 2020), n.breaks =12 ) +
  scale_fill_manual(values = c("#004D40", "#ff6600ff")) +
  geom_text(aes(label = count), 
            position=position_dodge(width = 0.9),
             vjust = -0.3, colour = "black", size=3.5) +
    geom_text(aes(label = total, y=0), 
            position=position_dodge(width = 0.9),
             vjust = -0.3, colour="white", size=3)
plotfall
ggsave("trappingeffort_fall.png")


plotspring <- trapeffort %>%
  filter(window=="spring") %>% 
  ggplot( aes(x=year, y=count, fill=trap)) +
  geom_col(stat = "identity", position = "dodge") +
  labs(y="count", title = "spring") +
  theme_classic() +
  scale_x_continuous(limits = c(2008, 2020), n.breaks =12 ) +
  # scale_y_continuous(limits = c(0, 130) ) +
  scale_fill_manual(values = c("#004D40", "lightblue")) +
  geom_text(aes(label = count), 
            position=position_dodge(width = 0.9),
             vjust = -0.3, colour = "black", size=3.5) +
    geom_text(aes(label = total, y=0), 
            position=position_dodge(width = 0.9),
             vjust = -0.3, colour="white", size=3)
plotspring
ggsave("trappingeffort_spring.png")


```

```{r}
ggpubr::ggarrange(plotfall,
          plotspring, 

          ncol=1)
ggsave("trappingeffort_2x30daywindows.png")
```



## same individuals per window?
Are the same individuals captured in the time window in yellow window of 19-days as before this window in 2009 and 2018 and 2019?

As you can see, we are constrained by a 19-day trapping window in the fall (by the start of trapping on 2013-06-03 and the end of trapping on 2009-06-23). Yet, some of the trapping effort is much bigger before the start date of trapping in other years (e.g., 2009 and 2018). Especially, we would miss individuals in fall 2018. Should we perhaps have two windows per season, accounting for dif-ferences in trapping design (timing of max trapping effort)? If so, would it matter if for some years we do not have trapping data at all for those time periods?
 
Sampling before the said window is not necessarily a problem. If animals marked prior to the sam-pling windows are alive when the sampling window begins they enter the study as newly marked individuals. So they become a part of the study.


compare missing ids before and after our chosen 2 x 30 day windows

ids per period/season/year

```{r fall}
year <- unique(trap$year)
season_select <- "fall"

#make empty dataframe to write loop results into
ids_fall_period <- data.frame(season=character(),
                       year=character(),
                       period=character(),
                       #period_19d=character(),
                       #period_after=character(),
                       ear_tag_id=character(),
                       sex=character(),
                       age=character())
#i=2009

for (i in year) {
  #r.seed <- run
  #set.seed(r.seed)
  run <- i
        
ids_trapped_fall <- trap %>% 
  mutate(period=if_else(between(julian_date, 155, 174), "select", 
                        if_else(julian_date < 155, "before", "after"))) %>%
  filter(year == i, season == season_select) %>%
  group_by(period, ear_tag_id) %>%
  slice(1)  %>%
  dplyr::select(season, period, year, ear_tag_id, sex, age)

  ids_trapped_fall <- as.data.frame(ids_trapped_fall)
  ids_fall_period <- rbind.data.frame(ids_fall_period, ids_trapped_fall)
  
}
```




```{r fall}


run <- unique(trap$year)
season_select <- "fall"

#make empty dataframe to write loop results into
missingids <- data.frame(season=character(),
                       year=character(),
                       period=character(),
                       n=numeric())
#i=2009
#run="2009"
#run <- i

for (i in run) {
  #r.seed <- run
  #set.seed(r.seed)
  #

ids_select <- trap %>%
  filter(season=="fall") %>%
  mutate(period=if_else(between(julian_date, 154, 184), "select", 
                        if_else(julian_date<154, "before", "after"))) %>%
  filter( year== i, period=="select")  %>%
  distinct(ear_tag_id) %>%
  arrange(ear_tag_id)#%>%
  #rename(select.id=ear_tag_id)

ids_before <- trap %>%
   filter(season=="fall") %>%
  mutate(period=if_else(between(julian_date, 154, 184), "select", 
                        if_else(julian_date<154, "before", "after"))) %>%
  filter( year==i, period=="before")  %>%
  distinct(ear_tag_id)%>%
  arrange(ear_tag_id) #%>%
 # rename(before.id=ear_tag_id)

ids_after <- trap %>%
   filter(season=="fall") %>%
  mutate(period=if_else(between(julian_date, 154, 184), "select", 
                        if_else(julian_date<154, "before", "after"))) %>%
  filter( year==i, period=="after")  %>%
  distinct(ear_tag_id)%>%
  arrange(ear_tag_id) 

# ids that are missing in the selected 19 day period that were captured before this window
idsmissing_select <- anti_join(ids_before, ids_select ) %>% #return all rows from x without a match in y.
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="idsmissing_select", 
         year=i, season=season_select)

# ids that are first trapped in the selected 19 day window
idstrapped_select <- anti_join(ids_select, ids_before) %>%
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="idstrapped_select", 
         year=i, season=season_select)

# ids that are missing in the selected 19 day period that were captured after this window
idsmissing_after <- anti_join(ids_after, ids_select) %>%
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="idsmissing_after", 
         year=i, season=season_select)

# total_trapped in window
idsfall <- trap %>%
  filter(season=="fall") %>%
  mutate(period=if_else(between(julian_date, 154, 184), "select", 
                        if_else(julian_date<154, "before", "after"))) %>%
  filter( year== i, period=="select") %>%
  summarize(n=n_distinct(ear_tag_id)) %>% 
  mutate(period="ndegus", 
         year=i, season=season_select)

# total_trapped 
idsall <- trap %>%
  filter(season=="fall",  year== i) %>%
  summarize(n=n_distinct(ear_tag_id)) %>% 
  mutate(period="total_ndegus", 
         year=i, season=season_select)

# trap nights in window
daysfall <- trap %>%
  filter(season=="fall") %>%
  mutate(period=if_else(between(julian_date, 154, 184), "select", 
                        if_else(julian_date<154, "before", "after"))) %>%
  filter( year== i, period=="select") %>%
  summarize(n=length(unique(date))) %>% 
  mutate(period="trapdays", 
         year=i, season=season_select)

# total trapnights per season
daysall <- trap %>%
  filter(season=="fall",  year== i) %>%
  summarize(n=length(unique(date))) %>% 
  mutate(period="total_trapdays", 
         year=i, season=season_select)


  ids_fall <- bind_rows(idsmissing_select,idstrapped_select, idsmissing_after, idsfall, idsall, daysfall, daysall  )
  missingids <- rbind.data.frame(missingids, ids_fall)
  
}

# in 2020 no degus were trapped at all. 
missingids$n[79] <-0
missingids$n[81] <-0
missingids$n[82] <-0

ids_fall_wide <- missingids %>%
  pivot_wider(
                             names_from = period, 
                             values_from = n)  %>%
  mutate(trap_effort=ndegus/total_ndegus, 
         month="jun") 



```


```{r spring}


run <- unique(trap$year)
season_select <- "spring"

#make empty dataframe to write loop results into
missingids.spring <- data.frame(season=character(),
                       year=character(),
                       period=character(),
                       n=numeric())
#i=2009
#run="2009"
#run <- i

for (i in run) {
  #r.seed <- run
  #set.seed(r.seed)
  #
        
ids_select <- trap %>%
  filter(season=="spring") %>%
  mutate(period=if_else(between(julian_date, 279, 298), "select", 
                        if_else(julian_date < 279, "before", "after"))) %>%
  filter( year== i, period=="select")  %>%
  distinct(ear_tag_id) %>%
  arrange(ear_tag_id)#%>%
  #rename(select.id=ear_tag_id)

ids_before <- trap %>%
   filter(season=="spring") %>%
  mutate(period=if_else(between(julian_date, 279, 298), "select", 
                        if_else(julian_date < 279, "before", "after"))) %>%
  filter( year==i, period=="before")  %>%
  distinct(ear_tag_id)%>%
  arrange(ear_tag_id) #%>%
 # rename(before.id=ear_tag_id)

ids_after <- trap %>%
   filter(season=="spring") %>%
  mutate(period=if_else(between(julian_date, 279, 298), "select", 
                        if_else(julian_date < 279, "before", "after"))) %>%
  filter( year==i, period=="after")  %>%
  distinct(ear_tag_id)%>%
  arrange(ear_tag_id) 

# ids that are missing in the selected 19 day period that were captured before this window
idsmissing_select <- anti_join(ids_before, ids_select ) %>% #return all rows from x without a match in y.
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="idsmissing_select", 
         year=i, season=season_select)

# ids that are first trapped in the selected 19 day window
idstrapped_select <- anti_join(ids_select, ids_before) %>%
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="idstrapped_select", 
         year=i, season=season_select)

# ids that are missing in the selected 19 day period that were captured after this window
idsmissing_after <- anti_join(ids_after, ids_select) %>%
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="idsmissing_after", 
         year=i, season=season_select)

# total_trapped in windowd
idsspring <- trap %>%
  filter(season=="spring") %>%
  mutate(period=if_else(between(julian_date, 279, 298), "select", 
                        if_else(julian_date < 279, "before", "after")))  %>%
  filter( year== i, period=="select") %>%
  summarize(n=n_distinct(ear_tag_id)) %>% 
  mutate(period="ndegus", 
         year=i, season=season_select)

# total_trapped 
idsall <- trap %>%
  filter(season=="spring",  year== i) %>%
  summarize(n=n_distinct(ear_tag_id)) %>% 
  mutate(period="total_ndegus", 
         year=i, season=season_select)

# trap nights in window
daysspring <- trap %>%
  filter(season=="spring") %>%
  mutate(period=if_else(between(julian_date, 279, 298), "select", 
                        if_else(julian_date<279, "before", "after"))) %>%
  filter( year== i, period=="select") %>%
  summarize(n=length(unique(date))) %>% 
  mutate(period="trapdays", 
         year=i, season=season_select)

# total trapnights per season
daysall <- trap %>%
  filter(season=="spring",  year== i) %>%
  summarize(n=length(unique(date))) %>% 
  mutate(period="total_trapdays", 
         year=i, season=season_select)

  ids_spring <- bind_rows(idsmissing_select,idstrapped_select, idsmissing_after, idsspring, idsall, daysspring, daysall)
  missingids.spring <- rbind.data.frame(missingids.spring, ids_spring)
  
}

# in 2020 no degus were trapped at all. 
missingids.spring$n[81] <-0
missingids.spring$n[82] <-0

ids_spring_wide <- missingids.spring %>%
  pivot_wider(names_from = period, 
              values_from = n) %>%
  mutate(trap_effort=ndegus/total_ndegus, 
         month="oct")


```

combine
```{r}

ids30days <- bind_rows(ids_fall_wide, 
                       ids_spring_wide) %>% 
  mutate(seasonid= )

write.csv(ids30days, "unique ids select 30day window.csv") 
```
have to change 2020 to 0 ids overall and trap effort of 0.0


## compares 30 day windows between fall and spring

```{r seasons}


run <- unique(trap$year)
season_select <- unique(trap$season)

#make empty dataframe to write loop results into
missingidsXseason <- data.frame(season=character(),
                       year=character(),
                       period=character(),
                       n=numeric())
#i=2009
#run="2009"
#run <- i

for (i in run) {
  #r.seed <- run
  #set.seed(r.seed)
  #
  
  ids_fall <- trap %>% 
    filter( year== i, between(julian_date, 154, 184)) %>%
  distinct(ear_tag_id) %>%
  arrange(ear_tag_id) 

    ids_spring <- trap %>% 
    filter( year== i, between(julian_date, 279, 309)) %>%
  distinct(ear_tag_id) %>%
  arrange(ear_tag_id)    

# ids that are missing in the selected 19 day period that were captured before this window
idsmissing_fall <- anti_join(ids_fall, ids_spring ) %>% #return all rows from x without a match in y.
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="ids.unique.in.fall", 
         year=i)

idsmissing_spring <- anti_join( ids_spring, ids_fall ) %>% #return all rows from x without a match in y.
  summarize(n=length(ear_tag_id)) %>% 
  mutate(period="ids.unique.in.spring", 
         year=i)


  idsXseason <- bind_rows(idsmissing_fall,idsmissing_spring  )
  missingidsXseason<- rbind.data.frame(missingidsXseason, idsXseason)
  
}

idsXseason <- missingidsXseason %>%
  pivot_wider(names_from = period, 
              values_from = n) 


```


# summary capture data

```{r}

capture_summary <- bind_rows(ids_fall_wide, 
                       ids_spring_wide) %>% 
  unite(seasonid, c("season", "month", "year"), remove = FALSE) %>%
  select(seasonid, season, year, month, 
         ndegus, total_ndegus, trap_effort,
         trapdays, total_trapdays, 
         idsmissing_select, idstrapped_select, idsmissing_after) %>%
  arrange(year) %>%
   mutate(seasonkey=1:24) %>%
  select(seasonkey, everything())

write.csv(capture_summary, "results/capture_summary.csv") 
```

We can add the environmental data to this file

# Mark recapture data

fall 
julian dates: 154 - 184
dates : 20xx-06-03 - 20xx-07-03

spring 
julian dates: 279 - 309
dates: 20xx-10-04 - 20xx-11-05


## certain death
add column: date_death -> only when in the notes the degu is persumed death, some deaths occured for second caputre so use original trap file then make selection (search on 'muer' and 'depr')

include a column where you use indicator variable (0 = no known death; 1 = known to have died just before the sampling occasion.

```{r add date deaths}

unique(trapdata$notes)

deaths<- read.csv("data/degus_deaths.csv") %>%
  dplyr::select(  ear_tag_id,  date_death)
#deaths$date <- as.Date(deaths$date)
deaths$ear_tag_id <- as.character(deaths$ear_tag_id)

markrecap_death <- trapdata %>% 
  left_join(deaths, by=c("ear_tag_id")) #%>%
  #mutate(known_death=if_else(is.na(date_death),0, 1))

check <- markrecap_death %>%
  filter(date_death!="NA")  %>%
  group_by(ear_tag_id) %>%
  slice(1)

# 10 missing perhaps because of subsequent captures so here i use unfiltered trapdata file. 

check <- deaths %>% left_join(markrecap) 
check <- check %>%  filter(julian_date=="NA")

death_degus_missing <- c("1501", "1554", "1688", "1644", "1709", 
  "3124", "3164", "6689", "6499", "6701")                   

check_trap <- trapdata %>% 
  filter(ear_tag_id %in% death_degus_missing)

```
## breeding status
add columns vulva -> 0=closed, 1=semi, 2=open
           

```{r}
unique(markrecap_death$breeding_status_1)

vulvasumm <- markrecap_death %>%
  group_by(breeding_status_1) %>% 
  tally()

markrecap_death$breeding_status_1[markrecap_death$breeding_status_1=="open!"] <- "open"

markrecap_breed <- markrecap_death %>%
  mutate(vulva=if_else(breeding_status_1=="close", 0, 
                       if_else(breeding_status_1=="open", 2, 
                              if_else(breeding_status_1=="semi", 1, NA))))
check <- markrecap_breed %>% filter(breeding_status_1=="semi")
check <- markrecap_breed %>% filter(breeding_status_1=="open")
check <- markrecap_breed %>% filter(breeding_status_1=="close")


```

 female_breeding <- for pregnant/lactating/ definitely pregnant = 1/ lact or pregn =4, ? = NS
 see breeding codes
preñada? ->	Se palpa preñez dudosa
x	-> No se palpa preñez

```{r}
unique(markrecap_breed$breeding_status_2)
length(unique(markrecap_breed$breeding_status_2))

breedsumm <- markrecap_breed %>%
  group_by(breeding_status_2) %>% 
  tally()

pregnant <- c("preñada", "Preñada", "no lactante, preñada"  )
lactating <- c("lactante",  "lactante, secresión sang.", 
               "lactante reciente", "lactante, no preñada", "lactante, X" )
postlactating <- c("post lactante",  "post lactante, no preñada" , "post lactante, X" , "postlactante", 
                    "postlactante, no preñada", "postlactante, X" )
lact.preg <- c("lactante, preñada", "post lactante, preñada", "preñada, con leche", "postlactante, preñada",
               "post lactante, preñada?", "post lactante, no preñada?!", 
                   "post lactante, no preñada?","postlactante, preñada?",
               "lactante, preñada?"  ,"lactante, no preñada?!", "post lactante?, preñada")
breedingstatus_unclear <- c("x", "P", "P?", "tapón mucoso", "?" , "preñada?", "preñada???",  "prenada?",
                            "lactante reciente?","post lactante?" , "lactante?, X", "Preñada?" )  
nonbreeding <- c("no lactante, no preñada!", "no preñada, no lactante" , "no preñada, no lactante, no preñada", 
                 "no lactante", "no lactante, no preñada"  )
na <- c("NA", NA)  

check <- length(pregnant)+length(lactating)+length(postlactating)+length(nonbreeding) +
  length(lact.preg)+length(na)+length(breedingstatus_unclear) # 1 missing

# create coding scheme
breeding <- c("pregnant", "lactating", "postlactating", "lact.preg", "breedingstatus_unclear", "nonbreeding", "na" )
code <- c(1, 2, 3, 4, 5, 0, NA)
description <- c("definitely pregnant", "definitely lactating", 
                 "definitely postlactating, female was pregnant and lactating this spring before this point",
                 "both pregnant and/or (post)lactating", 
                 "either breeding status with question marks or unknown abbreviation",
                 "female is not pregnant nor lactating", "NA" )
breeding_codes <- data.frame(breeding, code, description)
write.csv(breeding_codes, "data/breeding status 2 codes.csv")

# create new column
markrecap_breed1 <- markrecap_breed %>%
  mutate(breeding_status=if_else(breeding_status_2 %in% nonbreeding, 0, 
                       if_else(breeding_status_2 %in% pregnant, 1, 
                              if_else(breeding_status_2 %in% lactating, 2, 
                                      if_else(breeding_status_2 %in% postlactating, 3, 
                                              if_else(breeding_status_2 %in% lact.preg, 4,
                                                      if_else(breeding_status_2 %in% breedingstatus_unclear, 5, NA)))))))
check <- markrecap_breed1 %>% group_by(breeding_status_2) %>% slice(1)


```
 
## first capture only
filter on first capture only, order chronologically 
```{r}

markrecap <- markrecap_breed1 %>% 
  select(-orderkey) %>%
  arrange(year, season, date, julian_date) %>%
  filter(capture==1) %>%
  mutate(orderkey=1:n())

```


## season and sessionID
select season 
add column: seasonID -> "season-year-month", with NA's to rows that do not fall within the 30-day period
add column: sessionid -> fall 2009 =1 , spring 2009=2

```{r add seasonid}

markrecap_30daywindow <- markrecap %>%
  group_by(year) %>%
  mutate(window = if_else(between(julian_date, 154, 184), "jun", 
                          if_else(between(julian_date, 279, 309), "oct", NA))) %>%
  unite(seasonid, c("season", "window", "year"), remove = FALSE) %>%
  ungroup()

unique(markrecap_30daywindow$seasonid)

seasons <- markrecap_30daywindow %>%
  filter(window!="NA") %>% 
  group_by(year, seasonid) %>%
  tally() %>%
  arrange(year) %>% 
  ungroup() %>% 
  mutate(seasonkey=1:n()) %>%
  select(seasonkey, seasonid)

markrecap_key <- seasons %>% full_join(markrecap_30daywindow)
unique(markrecap_key$seasonkey)
unique(markrecap_key$seasonid)

trapkey <- seasons %>% 
  left_join(markrecap_30daywindow) %>%
  group_by(seasonkey) %>%
  slice(1) %>%
  select(seasonkey, seasonid, season, window, year) %>% 
  ungroup()

write.csv(trapkey, "./data/seasonkey.csv")
```

## clean up
select relevant columns and remove unnecessary columns
remove: orden_libreta, grid, burrow location.  

```{r}

str(markrecap_key)

trap <- markrecap_key %>%
  dplyr::select(orderkey,seasonkey, seasonid, season, window, year, date, julian_date, 
                trap_type, ear_tag_id, sex, radio_collar, weight, w_collar, 
                status, age, vulva, breeding_status, date_death, notes) %>%
  arrange(orderkey)
  #rename(id=ear_tag_id)

# grouping variables should be factor (sex, age)
trap$sex <- as.factor(trap$sex)
trap$age <- as.factor(trap$age)

str(trap)
```

##write file

```{r}
write.csv(trap, "data/degu_markrecapture.csv")
```

 



