---
title: "covariates"
author: "Annemarie van der Marel"
date: "2023-12-15"
output: pdf_document
---



Environmental variables
independent  and multiple estimates supports our hypothesis that drought affects vegetation production and hence food availability:

The values of NDVI indicate presence of green vegetation /indicator of vegetation greenness while LAI quantifies the amount of leaf material in a canopy. So, LAI is a better indicator of food availability.

The SMAP soil moisture is available only after 2015, but it clearly indicates the severe drought conditions in 2018. (perhaps as supplemental material)

Evapotranspiration contains both evaporation and transpiration, so I don't think I can use it as a proxy of soil moisture alone, but as an indicator of the effect of soil moisture.

The GPP values indicate  the total influx of carbon into an ecosystem through the photosynthetic fixation of CO2 and represents biomass production. So, indirectly support vegetation activity, partially contribute to E through transpiration.



# load libraries
```{r include=FALSE}
library(tidyverse)
library(paletteer)
library(ggrepel)
library(ggpubr)
library(dplyr)
library(gghighlight)
library(lubridate)
library(viridis)
library(patchwork)
```


# import data
```{r}

capture_sum <- read.csv("results/capture_summary.csv") %>%
  select(-X, -idsmissing_select, -idstrapped_select, -idsmissing_after) %>%
  dplyr::rename(trapsession=season, 
         window=month)

# seasonkey <- read.csv("./data/seasonkey.csv") %>%
#   select(-X) %>%
#   add_row(seasonkey=0, seasonid="spring_oct_2008", season="breeding", window="oct", year=2008, .before = 1) %>%
#   add_row(seasonkey=-1, seasonid="fall_jun_2008",season="nonbreeding", window="jun", year=2008, .before = 1)
  

seasonkey_month <- read.csv("./data/seasonkey_month_v2.csv") 

# environmental data
temp_raw <- read.csv("./data/monthly_data_Pudahuel.csv") 
temp_raw$precipitation_mm[temp_raw$precipitation_mm==-999.0] <- NA


enso_raw <- read.csv("./data/El Niño-Southern Oscillation.csv")

# envirownmental data _ MODIS
modis_day <- read.csv("results/MODIS_ET_GPP_LAI_dailyn_final.csv") %>% 
  dplyr::select(-X) %>% 
  dplyr::rename(year=Year,
        month=Month) 
 
modis_month <- read.csv("data/MODIS_monthly_data.csv") %>% # for monthly plots
  #dplyr::select(-X) %>% 
  dplyr::rename(year=Year,
        month=Month) 

```


# helper functions
```{r}
years <- c(2009:2020)
```


```{r}

library(colorblindr)
library(paletteer)
library(impressionist.colors)

paletteer::palettes_d_names
paletteer_d(`"Manu::Kea"`)
  paletteer_d(`"colorBlindness::paletteMartin"`)
paletteer_d(`"impressionist.colors::korenveld_onder_onweerslucht"`)
paletteer_d(`"impressionist.colors::les_nympheas"`)
paletteer_d(`"impressionist.colors::de_slaapkamer"`)


impressionist.colors::palette.summary()


colors<- paletteer_d(`"impressionist.colors::korenveld_onder_onweerslucht"`, 12)

colors24<- paletteer_d(`"impressionist.colors::korenveld_onder_onweerslucht"`, 24)
#colors <-  thematic::okabe_ito(12) "#E69F00" "#009E73" "#0072B2" "#CC79A7" "#999999" "#D55E00" "#F0E442" "#56B4E9"


 paletteer::scale_colour_paletteer_d("rcartocolor::Safe")
colorsXperiod <-paletteer::paletteer_d("rcartocolor::Safe")


cividis_colors <- viridis_pal(option = "cividis")(5)
print(cividis_colors)

print(viridis_pal(option = "D")(5))

 
```


# Weather covariates
## yearly temp and rainfall

```{r}
annual <- temp_raw %>%
  group_by(year) %>%
  dplyr::summarize(temp_avg=mean(avgT_C), 
            temp_max=mean(maxT_C),
            rain_mm=sum(precipitation_mm, na.rm = T),
            temp_cv=sd(avgT_C, na.rm = T) / mean(avgT_C, na.rm = T), 
            rain_cv=sd(precipitation_mm, na.rm = T)/mean(precipitation_mm, na.rm = T))
head(annual)
#annual$year <- as.integer(annual$year)
```

Percent Decline=( Original Value -New Value)/Original value *100

 
```{r}
# average rain_sum 1976-2008
meanprior <- annual %>% 
  filter(year<2009) %>% 
  summarize(mean_rain=mean(rain_mm))
  
# average rainfall 2009-2019
meanstudy<- annual %>% 
  filter(between(year, 2009, 2019)) %>% 
  summarize(mean_rain=mean(rain_mm))

percdecline <- (meanprior$mean_rain-meanstudy$mean_rain)/meanprior$mean_rain *100
```


## monthly temperature and rainfall
```{r}
month_num <- c(1:12)
month <- c("jan", "feb", "mar", "apr", "may", "jun", 
           "jul", "aug", "sep", "oct", "nov", "dec")

months <- data.frame(month, month_num)
```

## plots monthly average study period

```{r temp+rain 2009-2020}

range(temp_raw$avgT_C)

plot_temp_month <-  temp_raw %>%
  filter(year>2008) %>%
  ggplot(aes(x=month, y=avgT_C, group=as.factor(year), colour=as.factor(year))) +   geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =25),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  geom_line() +   geom_point() +
  theme_classic() +
  #facet_wrap(~year) +
  scale_y_continuous(limits=c(0, 25)) +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
  geom_hline(yintercept = mean(temp_raw$avgT_C), linetype="dotted") + #mean(temp_raw$avgT_C)
  #scale_color_continuous("PuBuGn") +
  labs(x="Month",y="T-avg"~(degree*C), subtitle = expression('Air Temperature'), colour="year") +  
  scale_color_manual(values = colorsXyear, name="Year") +
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray") +
   geom_hline(yintercept = mean(annual$temp_avg), linetype="dotted")
plot_temp_month   

range(temp_raw$maxT_C)
plot_tempmax_month <-  temp_raw %>%
  filter(year>2008) %>%
  ggplot(aes(x=month, y=maxT_C, group=as.factor(year), colour=as.factor(year))) +   geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =35),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  geom_line() +   geom_point() +
  theme_classic() +
  #facet_wrap(~year) +
  #scale_y_continuous(limits=c(0, 25)) +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
  geom_hline(yintercept = mean(temp_raw$maxT_C), linetype="dotted") + #mean(temp_raw$avgT_C)
  #scale_color_continuous("PuBuGn") +
  labs(x="Month", y="T max"~(degree*C), subtitle ="Max Temperature", colour="year") +  
  scale_color_manual(values = colorsXyear, name="Year") 
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray")
plot_tempmax_month 

range(temp_raw$precipitation_mm, na.rm = T)

plot_rain_month <-  temp_raw %>%
  filter(year>2008) %>%
  ggplot( aes(x=month, y=precipitation_mm, group=as.factor(year), colour=as.factor(year))) +
    geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =100),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  geom_line() +   geom_point() +
  theme_classic() +
  #facet_wrap(~year) +
  labs(x="Month", y="P (mm)", subtitle ="Precipitation", colour="year") +
  geom_hline(yintercept = mean(temp_raw$precipitation_mm, na.rm = T), linetype="dotted") + # mean of the total amount of yearly rainfall from he study period
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
  scale_y_continuous(limits=c(0, 100)) +
  scale_color_manual(values = colorsXyear, name="Year") 
  #facet_wrap(~year) +
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray")
plot_rain_month


```
##compare long-term vs study period
```{r}
covars <- read.csv("./data/covariates_select_scaled_v5_exclude-temp-max.csv") %>% 
  filter(year2!=2020)

covars$season2 <- factor(covars$season2, levels = c("wet", "dry") )

covarssum <- covars %>% 
  dplyr::group_by(year2, season2) %>% 
  summarize(min=min(rain_sum),
            max=max(rain_sum))

ggplot(covars, aes(x=year2, y=rain_sum)) +
         geom_point() +
         facet_grid(~season)

obsmean_rain_sum <- mean(covars$rain_sum)
obssd_rain_sum <- sd(covars$rain_sum)

obsmean_rain_cv_lag <- mean(covars$rain_cv_lag)
obssd_rain_cv_lag <- sd(covars$rain_cv_lag)

# environemntal covars
temp_raw <- read.csv("./data/monthly_data_Pudahuel.csv") %>% 
  filter(year!="2020")
temp_raw$precipitation_mm[temp_raw$precipitation_mm==-999.0] <- NA

2019-1976



annual <- temp_raw %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(temp_avg=mean(avgT_C), 
            temp_mMax=mean(maxT_C),
            rain_mm=sum(precipitation_mm, na.rm = T))

rainavg43years <- mean(annual$rain_mm, na.rm = T)
round(rainavg43years)

study <- temp_raw %>%
  filter(year>2008) %>% 
  dplyr::group_by(year) %>%
  dplyr::summarize(temp_avg=mean(avgT_C), 
            temp_mMax=mean(maxT_C),
            rain_mm=sum(precipitation_mm, na.rm = T))
rainavg12years <- round(mean(study$rain_mm, na.rm=T))
```
The long-term yearly average of cumulative amount of rainfall between 1976 and 2019 was 247 mm. The rainfall that fell between 2009 and 2019 that correspond to our data set was 153 mm. 

## select weather variables
```{r}
head(temp_raw)

nonbreeding <- c(1, 2, 3,4, 5, 10, 11, 12) #oct-march
breeding <- c(6:9) #april-sept

seasonkey_month$year<- as.integer(seasonkey_month$year)


temp <- seasonkey_month %>% 
  filter(year!="2021") %>% 
  left_join(temp_raw) %>%
  dplyr::group_by(season, window, seasonkey, seasonid) %>%
 # arrange( year, season) %>%
  dplyr::summarise(temp_avg=mean(avgT_C, na.rm = T),
            temp_cv=sd(avgT_C, na.rm = T) / mean(avgT_C, na.rm = T), 
            temp_max=max(maxT_C, na.rm = T),
            rain_sum=sum(precipitation_mm, na.rm = T), 
            rain_cv=sd(precipitation_mm, na.rm = T)/mean(precipitation_mm, na.rm = T)) %>%
  dplyr::arrange(seasonkey) %>% 
  ungroup() 


temp$rain_cv[is.na(temp$rain_cv)]<-0



```

## lagged weather

Select appropriate weather variables for each season (lagged etc)


1.	rainfall data from Jun-Sept year n to predict survival/recruitment in June-Sept year n+1
2.	rainfall data from Oct-May year n to predict survival/recruitment June-Sept year n
It depends on how much effect we think rainfall will have: does the rainfall from a year ago influ-ence current food availability or is it more of an immediate effect, where rainfall in April/May in-fluence food availability? And I'm not sure which one to choose. 
This question arose because we changed the seasons for which we combined the environmental covariates from Apr-Sept to Jun-Sep, which meant there is a bit more rainfall during the Oct-May interval. In my email from last week, I attached an image of the sum of monthly rainfall across the 2009-2020 interval. 

Answer from Praveena: The impact of rainfall on vegetation depends on the type of vegetation.  If it's herbaceous plants (important for Degus), the effect is immediate, but plants with larger roots or trees can have a noticeable lag effect , which I guess is not much important to Degus. I think we should give more preference to the immediate growing season than the previous years or next years. Severe drought can definitely affect plant  density, leaf area next year due to impact on its growth in the first year affecting seed pool or bud formation for the next year.

Answer from Loren: The long-term effect may depend on how cumulative rainfall if the seed base depends on this. If it is impossible to include multiple periods, then I suggest going with the immediate effect. We might need a more in depth understanding of the long=term effects of rainfall on seed germina-tion, plant growth, etc.


```{r}

glimpse(temp)

weatherlag <- temp %>% 
  mutate( # focus on immediate effects
          rain_sum_lag=if_else(season=="breeding", lag(rain_sum, 1), lag(rain_sum, 1)),
          rain_cv_lag=if_else(season=="breeding", lag(rain_cv, 1), lag(rain_cv, 1)) ) # ,
    
    # for lag rain, breeding season is former wet season, dry season is former wet season
          # rain_sum_lag2=if_else(season=="breeding", lag(rain_sum, 2), lag(rain_sum, 1)),
          # rain_cv_lag2=if_else(season=="breeding", lag(rain_cv, 2), lag(rain_cv, 1))) 


```



# Environmental covariates

##study period
```{r}
modis_study <- modis_month %>% 
  filter(year %in% years)


```


## LAI: Leaf Area Index

```{r}
glimpse(modis_day)

  # plots
plot_lai_month <-  modis_study %>%
  ggplot( aes(x=month, y=LAI_avg, group=as.factor(year), colour=as.factor(year))) +
      geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =3),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  geom_line() +   geom_point( )+
  theme_classic() +
  scale_color_manual(values = colorsXyear, name="Year") +
  #facet_wrap(~year) +
  labs(x="Month", y=expression(LAI-avg~(m^2)), subtitle ="Leaf Area Index", colour="year") +
  geom_hline(yintercept = mean(modis_study$LAI_avg, na.rm=T), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
  scale_y_continuous(limits=c(0, 3)) #max(modis_study$LAI_avg)
 
  #geom_segment(x = 6, xend=9.9, y =6, yend=5.6, colour="lightgray", alpha=0.7)
plot_lai_month


```




## ET: Evapotranspiration

```{r}
max(modis_study$ET_sum)

plot_et_sum <-  modis_study %>%
  ggplot( aes(x=month, y=ET_sum, group=as.factor(year), colour=as.factor(year))) +
   geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =71),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
    geom_line() +   geom_point() +
  scale_color_manual(values = colorsXyear, name="Year") +
  theme_classic() +
  #facet_wrap(~year) +
  labs(x="Month", y="ET (mm)", subtitle ="Evapotranspiration", colour="year") +
  geom_hline(yintercept = mean(modis_study$ET_sum), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) #+
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray")
plot_et_sum

max(modis_study$ET_avg)

  plot_et_avg <-  modis_study %>%
   ggplot( aes(x=month, y=ET_avg, group=as.factor(year), colour=as.factor(year))) +
   geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =3),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  scale_color_manual(values = colorsXyear, name="Year") +
  geom_line() +   geom_point() +
  #scale_color_manual(values = colors) +
  theme_classic() +
  #facet_wrap(~year) +
  labs(x="Month", y="ET-avg (mm)", subtitle ="Evapotranspiration", colour="year") +
  geom_hline(yintercept = mean(modis_study$ET_avg), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
  geom_vline(xintercept = 5.9, colour="lightgray") +
  geom_vline(xintercept = 9.9, colour="lightgray")
plot_et_avg

ggarrange(plot_et_avg, plot_et_sum)
```
## GPP: Gross Primary Production

```{r}
max(modis_study$GPP_sum)

plot_gpp_sum <-   ggplot( modis_study, 
                          aes(x=month, y=GPP_sum, group=as.factor(year), colour=as.factor(year))) +
    geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = 0, ymax =150),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  geom_line() +   geom_point() +
  theme_classic() +
  scale_color_manual(values = colorsXyear, name="Year") +
  #facet_wrap(~year) +
  labs(x="Month", y=expression(GPP~(gC/m^2)), subtitle ="Gross Primary Production", colour="year") +
  geom_hline(yintercept = mean(modis_study$GPP_sum), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) 
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray")
plot_gpp_sum

plot_gpp_avg<-  ggplot( modis_study, 
                          aes(x=month, y=GPP_avg, group=as.factor(year), colour=as.factor(year))) +
  geom_line() +   geom_point() +
  theme_classic() +
  scale_color_manual(values = colors) +
  #facet_wrap(~year) +
  labs(x="Month", y=expression("GPP avg"~(gC/m^2)), subtitle ="Gross Primary Production", colour="year") +
  geom_hline(yintercept = mean(modis_study$GPP_avg), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) 
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray")
plot_gpp_avg

ggarrange(plot_gpp_sum, plot_gpp_avg)

```



## ONI
El Niño-Southern Oscillation
Regional temperature and precipitation oscillations caused by El Nino. “El Niño conditions occur when the average sea-surface temperature of three consecutive months is 0.5°C above average temperature and La Niña conditions occur when the average sea-surface temperature is 0.5°C below average temperature. We obtained the ONI value from the National Oceanographic and Atmospheric Agency Climate Prediction Center” https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php 

```{r}

# did it in excel instead, see
oni<- seasonkey_month %>%
  left_join(enso_raw, by=c("year", "month")) %>%
  dplyr::filter(oni!="NA") %>% 
  dplyr::group_by(season, window, seasonkey, seasonid) %>%
  dplyr::summarize(oni_avg=mean(oni)) %>% 
  dplyr::arrange(seasonkey) 

range(enso_raw$oni)

plot_oni_month <-  enso_raw %>%
  filter(year%in%years) %>% 
  group_by(year, month) %>%
  #summarize(mean_oni=mean(oni)) %>% 
  #filter(year>2016) %
  ggplot( aes(x=month, y=oni, group=as.factor(year), colour=as.factor(year))) +
    geom_rect( aes( xmin = 5.9,  xmax = 9.9,
      ymin = -2, ymax =3),
    fill = alpha("lightgrey", 0.2), color="lightgrey") +
  geom_line() +   geom_point() +
  theme_classic() +
  scale_color_manual(values = colorsXyear, name="Year") +
  #facet_wrap(~year) +
  labs(x="Month", y="ONI", subtitle ="El Niño Index", colour="year") +
  geom_hline(yintercept = mean(enso_raw$oni), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) 
  #scale_y_continuous(limits=c(0, 0.05)) +
  # geom_vline(xintercept = 5.9, colour="lightgray") +
  # geom_vline(xintercept = 9.9, colour="lightgray")
plot_oni_month

```


# Plots Study Period Monthly

```{r}
plot_month_all <- ggpubr::ggarrange(plot_temp_month + rremove("xlab"), 
                                    plot_lai_month+ rremove("xlab"), 
                                    plot_rain_month+ rremove("xlab"),  
                                    plot_et_sum+ rremove("xlab"),
                                    plot_tempmax_month+ rremove("xlab"), 
                                    plot_gpp_sum, plot_oni_month,
                         nrow=4, ncol=2, align = "v",
                         common.legend = TRUE, legend = "bottom")
plot_month_all

ggsave( "results/environmental covars X month_allv4.pdf", width = 180, height = 200, units = "mm", dpi=300)
```





# Plots long-term covariates

https://davidmathlogic.com/colorblind/#%23D81B60-%231E88E5-%23FFC107-%23004D40 
```{r annual}
head(temp_raw)
temp_raw$precipitation_mm[temp_raw$precipitation_mm==-999.0] <- NA

annual <- temp_raw %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(temp_avg=round(mean(avgT_C), 2), 
            temp_max=round(max(maxT_C, na.rm = T), 2),
            #temp_cv=formatC(sd(avgT_C, na.rm = T) / mean(avgT_C, na.rm = T),format = "f", flag="0"),
            temp_cv=round(sd(avgT_C, na.rm = T) / mean(avgT_C, na.rm = T),3),
            rain_mm=sum(precipitation_mm, na.rm = T),
            rain_cv=round(sd(precipitation_mm, na.rm = T)/mean(precipitation_mm, na.rm = T), 3))

plot_temp_year <- ggplot(annual, aes(x=year, y=temp_avg)) +
  geom_line(colour="#fd8d3c") +  
  geom_point(fill="#fd8d3c", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year", y=expression('T avg'~(degree*C)), subtitle ='Air Temperature')+
  geom_hline(yintercept = mean(annual$temp_avg), linetype="dotted")
plot_temp_year

plot_temp_cv_year <- ggplot(annual, aes(x=year, y=temp_cv)) +
  geom_line(colour="#fed976") +   
  geom_point(fill="#fed976", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year", y="T CV", subtitle = "CV Temperature")+
  geom_hline(yintercept = mean(annual$temp_cv), linetype="dotted")
plot_temp_cv_year

plot_temp_max_year <- ggplot(annual, aes(x=year, y=temp_max)) +
  geom_line(colour="#f03b20") +   
  geom_point(fill="#f03b20", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year",  y=expression('T max'~(degree*C)), subtitle ="Max Temperature")+
  geom_hline(yintercept = mean(annual$temp_max), linetype="dotted")
plot_temp_max_year

plot_rain_year <- ggplot(annual, aes(x=year, y=rain_mm)) +
  geom_line(color="#1E88E5") +  
  geom_point(fill="#1E88E5", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year", y="P (mm)", subtitle =  "Precipitation") +
  geom_hline(yintercept = mean(annual$rain_mm), linetype="dotted") +
  scale_y_continuous(limits=c(0, 700))
plot_rain_year

range(annual$rain_cv)
plot_raincv_year <- ggplot(annual, aes(x=year, y=rain_cv)) +
  geom_line(color="#9ecae1") +  
  geom_point(fill="#9ecae1", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year", y="P CV", subtitle =  "CV Precipitation") +
  geom_hline(yintercept = mean(annual$rain_cv), linetype="dotted") +
  scale_y_continuous(limits=c(0, 2.5))
plot_raincv_year

plot_lai_year <- modis_day %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(lai_avg=mean(LAI)) %>% 
  ggplot( aes(x=year, y=lai_avg)) +
  geom_line(color="#a1d99b") +  
  geom_point(fill="#a1d99b", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year", y=expression("LAI"~(m^2)), subtitle =  "Leaf Area Index") +
  geom_hline(yintercept = mean(modis_day$LAI), linetype="dotted") +
  #scale_y_continuous(limits=c(0, 0.5)) +
  scale_x_continuous(limits=c(2000, 2020))
plot_lai_year 

plot_gpp_year <- modis_day %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(gpp_avg=mean(GPP)) %>% 
  ggplot( aes(x=year, y=gpp_avg)) +
  geom_line(color="#006d2c") +  
  geom_point(fill="#006d2c", shape = 21, color="black",  size=2 ) +
  theme_classic() +
  labs(x="Year", y=expression('GPP'~(gCm^2)), subtitle="Gross Primary Production") +
  geom_hline(yintercept = mean(modis_day$GPP), linetype="dotted") +
  #scale_y_continuous(limits=c(0, 0.5)) +
  scale_x_continuous(limits=c(2000, 2020))
plot_gpp_year 

plot_et_year <- modis_day %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(et_avg=mean(ET)) %>% 
  ggplot( aes(x=year, y=et_avg)) +
  geom_line(color="#D81B60") +  
  geom_point(shape = 21, color="black", fill="#D81B60", size=2) +
  theme_classic() +
  labs(x="Year", y="ET (mm)",subtitle ="Evapotranspiration") +
  geom_hline(yintercept = mean(modis_day$ET), linetype="dotted") +
  #scale_y_continuous(limits=c(0, 0.5)) +
  scale_x_continuous(limits=c(2000, 2020))
plot_et_year 


plot_oni_year <- enso_raw %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(oni_avg=mean(oni)) %>% 
  ggplot( aes(x=year, y=oni_avg)) +
  geom_line(color="#bdbdbd") +  
  geom_point( shape = 21, color="black", fill="#bdbdbd", size=2 ) +
  theme_classic() +
  labs(x="Year", y= "ONI", subtitle="El Niño index") +
  geom_hline(yintercept = mean(enso_raw$oni, na.rm=T), linetype="dotted") +
  #scale_y_continuous(limits=c(0, 0.5)) +
  scale_x_continuous(limits=c(2000, 2020)) 
  
plot_oni_year

```


```{r}

ggarrange(plot_temp_year, plot_lai_year, 
          plot_temp_max_year, plot_gpp_year,
          plot_temp_cv_year, plot_et_year,
          plot_rain_year, plot_oni_year,
          plot_raincv_year, 
          nrow=5, ncol=2, 
          align="hv", widths = c(1.5,1))

ggsave("./results/longterm_envir-covars_v10.pdf", width = 180, height=240, units = "mm", dpi=300)

```





# Select Covariates


## combine all covars
```{r}
# By day
modis_day$year <- as.factor(modis_day$year)

modis <- seasonkey_month %>%
  left_join(modis_day) 

## join to weather variables
modisXseason <- modis %>%
  dplyr::group_by(season, seasonkey, seasonid) %>%
    dplyr::summarize(gpp_avg=mean(GPP, na.rm = T), 
            gpp_sum = sum(GPP, na.rm = T),
            et_avg=mean(ET, na.rm=T),
            et_sum = sum(ET, na.rm = T),
            lai_avg = mean(LAI, na.rm=T)) %>% 
  dplyr::arrange(seasonkey)



# by month
modis_month$year<- as.factor(modis_month$year)
modism <- seasonkey_month %>%
  left_join(modis_month) 

## join to weather variables
modismXseason <- modism %>%
  dplyr::group_by(season, seasonkey, seasonid) %>%
    dplyr::summarize(gpp_avg=mean(GPP_avg, na.rm = T), 
            gpp_sum = sum(GPP_sum, na.rm = T),
            et_avg=mean(ET_avg, na.rm=T),
            et_sum = sum(ET_sum, na.rm = T),
            lai_avg = mean(LAI_avg, na.rm=T)) %>% 
  arrange(seasonkey)




# combine all weather variables
weather_modis <- weatherlag %>% 
  left_join(modisXseason, by = c( "seasonkey", "season", "seasonid")) %>% 
  left_join(oni) %>% 
  filter(seasonkey>0)
glimpse(weather_modis)


write.csv(weather_modis, "data/all environmental covariates_20250225.csv")

```



#Pairwise correlations
http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software

my general rule is to not use variables with r>=0.5.



## select sum GPP and ET as we also have sum P

```{r}
covariates <- weather_modis %>% 
  dplyr::select(-gpp_avg, -et_avg)

# select variables to include in correlation matrix
df <- covariates[, 5:15]

#create matrix
res <- round(cor(df),2)
write.csv(res, "./results/covariate correlation_v10.csv")


# visalise matrix
pairs(df)
psych::pairs.panels(df) # Generate scatterplots with Pearson correlations

jpeg(file="./results/correlation matrix_v10.jpeg")
corrplot::corrplot(round(cor(df),2), type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
dev.off()

#PerformanceAnalytics::chart.Correlation(my_data, histogram=TRUE, pch=19)

#library(Hmisc)
#res2<-Hmisc::rcorr(as.matrix(covariates[,5:15]))

#corrplot::corrplot(res2$r, type="upper", order="hclust", 
#         p.mat = res2$P, sig.level = 0.01, insig = "blank")


```

### R

 We have to exclude max temperature as it still has r2>0.5 with rain_sumlag and rain_cv_lag.  

```{r}

# select variables to include in correlation matrix
select_covariates<- covariates %>%
  #dplyr::select(-temp_avg) # still ones with r > 0.5
  #dplyr::select(-temp_avg,  -temp_max) # 
  #dplyr::select(-temp_avg,  -rain_sum_lag, - temp_max) # Tmax r>0.5 with 3 other variables
  dplyr::select(-temp_avg,  -rain_sum_lag, - temp_max, -et_sum) # Tmax r>0.5 with 3 other variables

dfs <- select_covariates[,5:11]
ress <- round(cor(dfs),2)
psych::pairs.panels(df1) # Generate scatterplots with Pearson correlations



```

If we use sum, then remove Temp_avg, temp_max, rain_sum_lag, and et_sum


###VIF
Exclude variables with VIF values > 3 or >10? 



```{r}
# vif
Z <- cbind( covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            covariates$et_sum, 
            covariates$gpp_sum,
            covariates$oni_avg)
colnames(Z) <- c( "T", "Tcv","Tmax", "P","Plag", "Pcv","Pcvl", "lai",  "et","gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: T


# Relationships between explanatory variables.
Z <- cbind( #covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            covariates$et_sum, 
            covariates$gpp_sum,
            covariates$oni_avg)
colnames(Z) <- c(  "Tcv","Tmax", "rain","rainlag", "raincv","raincvl", "lai",  "et","gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: et


# Relationships between explanatory variables.
Z <- cbind( #covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            #covariates$et_avg, 
            covariates$gpp_sum,
            covariates$oni_avg)
colnames(Z) <- c(  "Tcv","Tmax", "rain","rainlag", "raincv","raincvl", "lai",  "gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: tmax

Z <- cbind( #covariates$temp_avg, 
            covariates$temp_cv,
            #covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            #covariates$et_avg, 
            covariates$gpp_sum,
            covariates$oni_avg)
colnames(Z) <- c(  "Tcv", "rain","rainlag", "raincv","raincvl","lai", "gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: 


# varaibales all with vif<3, exclude temp_avg, tempmax, et_sum,


```

## select avg GPP and ET so not to change to much 

```{r}
covariates <- weather_modis %>% 
  dplyr::select(-gpp_sum, -et_sum)

# select variables to include in correlation matrix
df <- covariates[, 5:15]

#create matrix
res <- round(cor(df),2)
write.csv(res, "./results/covariate correlation_v10.csv")


# visalise matrix
pairs(df)
psych::pairs.panels(df) # Generate scatterplots with Pearson correlations

jpeg(file="./results/correlation matrix_v10.jpeg")
corrplot::corrplot(round(cor(df),2), type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
dev.off()

#PerformanceAnalytics::chart.Correlation(my_data, histogram=TRUE, pch=19)

#library(Hmisc)
#res2<-Hmisc::rcorr(as.matrix(covariates[,5:15]))

#corrplot::corrplot(res2$r, type="upper", order="hclust", 
#         p.mat = res2$P, sig.level = 0.01, insig = "blank")


```
### R

 We have to exclude max temperature as it still has r2>0.5 with rain_sumlag and rain_cv_lag.  

```{r}

# select variables to include in correlation matrix
select_covariates<- covariates %>%
  #dplyr::select(-temp_avg) # still ones with r > 0.5
  #dplyr::select(-temp_avg,  -et_avg) # 
  #dplyr::select(-temp_avg,  -et_avg, - lai_avg) # Tmax r>0.5 with 3 other variables
  dplyr::select(-temp_avg,-temp_max,  -et_avg, - lai_avg) # Tmax r>0.5 with 3 other variables

df1 <- select_covariates[,5:11]
res1 <- round(cor(df1),2)
psych::pairs.panels(df1) # Generate scatterplots with Pearson correlations



```


### VIF
```{r}
# vif
Z <- cbind( covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            covariates$et_avg, 
            covariates$gpp_avg,
            covariates$oni_avg)
colnames(Z) <- c( "T", "Tcv","Tmax", "rain","rainlag", "raincv","raincvl", "lai",  "et","gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: T


# Relationships between explanatory variables.
Z <- cbind( #covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            covariates$et_avg, 
            covariates$gpp_avg,
            covariates$oni_avg)
colnames(Z) <- c(  "Tcv","Tmax", "rain","rainlag", "raincv","raincvl", "lai",  "et","gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: et


# Relationships between explanatory variables.
Z <- cbind( #covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            covariates$lai_avg, 
            #covariates$et_avg, 
            covariates$gpp_avg,
            covariates$oni_avg)
colnames(Z) <- c(  "Tcv","Tmax", "rain","rainlag", "raincv","raincvl", "lai",  "gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: lai

Z <- cbind( #covariates$temp_avg, 
            covariates$temp_cv,
            covariates$temp_max,
            covariates$rain_sum,
            covariates$rain_sum_lag, 
            covariates$rain_cv,
            covariates$rain_cv_lag, 
            #covariates$lai_avg, 
            #covariates$et_avg, 
            covariates$gpp_avg,
            covariates$oni_avg)
colnames(Z) <- c(  "Tcv","Tmax", "rain","rainlag", "raincv","raincvl", "gpp", "oni")
dfz = data.frame(Z) # Data Frame with predictor variables
usdm::vif(dfz) # VIF > 10: 


# varaibales all with vif<3, exclude temp_avg, et_avg, lai_avg

```









# standardize and save select covariates

If we use sum, then remove Temp_avg, temp_max, rain_sum_lag, and et_sum

```{r}
#center on mean and scale on SD
# scale_covariates_sum <- select_covariates %>% 
#   mutate(temp_cv2=scale(temp_cv), 
#          rain_sum2=scale(rain_sum),
#          rain_cv2=scale(rain_cv),
#          gpp_sum2=scale(gpp_sum),
#          lai_avg2=scale(lai_avg),
#          oni_avg2=scale(oni_avg),
#          rain_cv_lag2=scale(rain_cv_lag), 
#          season2=if_else(season=="breeding", "wet", "dry")
#         ) %>% 
#   separate(seasonid, into = c("season1", "window1", "year2"), remove = F) %>% 
#   select(seasonkey, seasonid, season, season1, season2, window, year2, everything(), -window1)
# 
# 
# write.csv(scale_covariates_sum, "./data/covariates_select_scaled_v10_sum.csv")

```

If we use avg then remove -temp_avg,-temp_max,  -et_avg, - lai_avg

```{r}
#center on mean and scale on SD
scale_covariates_avg <- select_covariates %>% 
  mutate(temp_cv2=scale(temp_cv), 
         rain_sum2=scale(rain_sum),
         rain_cv2=scale(rain_cv),
         gpp_avg2=scale(gpp_avg),
         oni_avg2=scale(oni_avg),
         rain_sum_lag2=scale(rain_sum_lag), 
         rain_cv_lag2=scale(rain_cv_lag), 
         season2=if_else(season=="breeding", "wet", "dry")
        ) %>% 
  separate(seasonid, into = c("season1", "window1", "year2"), sep="_", remove = FALSE) %>% 
  select(seasonkey, seasonid, season, season1, season2, window, year2, everything(), -window1)


write.csv(scale_covariates_avg, "./data/covariates_select_scaled_v10_avg.csv")

```



# Plots study period
##plots highlight 2019

```{r}
colorsXyear <- paletteer::scale_colour_paletteer_d("rcartocolor::Safe")
colorsXyear <-paletteer::paletteer_d("rcartocolor::Safe")
```


```{r GPP}

plot_gpp_hl<-  ggplot( modis_study, 
                          aes(x=month, y=GPP_avg, group=as.factor(year), colour=as.factor(year))) +
  #   geom_rect( aes( xmin = 5.9,  xmax = 9.9, 
  #     ymin = 0, ymax = 5),
  #   fill = "lightgrey", color="lightgrey", alpha=0.2) +
  geom_line(size=1) +   geom_point(size=3) +
  scale_color_manual(values = colorsXyear) +
   gghighlight(year=="2019",
              # preserve colour and modify alpha instead
              unhighlighted_params = list(colour = NULL, alpha = 0.3, size=1))+
  theme_degu() +
  theme(legend.position="right") +
  #facet_wrap(~year) +
  labs(x="Month", y=expression("GPP avg"~(gCm^2)),  colour="year") + #subtitle ="Gross Primary Production",
  geom_hline(yintercept = mean(modis_study$GPP_avg), linetype="dotted") +
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
   geom_vline(xintercept = 5.9,  linetype="dashed", colour="gray") + #colour="lightgray",
  geom_vline(xintercept = 9.9, linetype="dashed", colour="gray")
plot_gpp_hl


```
 



```{r rain sum}
temp_study <- temp_raw %>% 
 filter(year %in% years)
temp_study$year <- as.factor(temp_study$year)


plot_rain_hl <-  temp_study %>%
  ggplot( aes(x=month, y=precipitation_mm, group=as.factor(year), colour=as.factor(year))) +
  geom_line(size=1) +   geom_point(size=3) +
  scale_color_manual(values=colorsXyear) +
  gghighlight(year=="2019", #label_key = year,
              # preserve colour and modify alpha instead
              unhighlighted_params = list(colour = NULL, alpha = 0.3, size=1))+
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
    scale_y_continuous(limits=c(0, 100)) +
  labs(x="Month", y="P (mm)", colour="year") +
  geom_hline(yintercept = mean(annual$rain_mm, na.rm=T), linetype="dotted") + # mean of the total amount of yearly rainfall from 1976-2020
  geom_hline(yintercept = mean(temp_study$precipitation_mm, na.rm=T), linetype="dotted") + # mean of the total amount of yearly rainfall from 1976-2020
  geom_vline(xintercept = 5.9,  linetype="dashed", colour="gray") + #colour="lightgray",
  geom_vline(xintercept = 9.9, linetype="dashed", colour="gray") + #colour="lightgray")
   # geom_text_repel(data = labels_df, aes(x=month, y=precipitation_mm, label = year), 
   #            nudge_x = 0.0, nudge_y = 0.001, size = 3, color = "black",
   #             max.overlaps = getOption("ggrepel.max.overlaps", default = 15)) +
   theme_degu() +
   theme(legend.position = "none")
plot_rain_hl
```



```{r oni}
enso_study <- enso_raw %>% 
   filter(year %in% years)

enso_study$year <- as.factor(enso_study$year)


plot_oni_hl <-  enso_study %>%
  ggplot( aes(x=month, y=oni, group=as.factor(year), colour=as.factor(year))) +
  geom_line(size=1) +   geom_point(size=3) +

  scale_color_manual(values=colorsXyear) +
  gghighlight(year=="2019", #label_key = year,
              # preserve colour and modify alpha instead
              unhighlighted_params = list(colour = NULL, alpha = 0.3, size=1))+
  scale_x_continuous(limits=c(1,12), breaks = 1:12) +
  
  labs(x="Month",  y ="ONI", colour="year") +
  
  geom_hline(yintercept = mean(enso_raw$oni), linetype="dotted") +
  geom_vline(xintercept = 5.9,  linetype="dashed", colour="gray") + #colour="lightgray",
  geom_vline(xintercept = 9.9, linetype="dashed", colour="gray") + #colour="lightgray")
   # geom_text_repel(data = labels_df, aes(x=month, y=oni, label = year), 
   #            nudge_x = 0.0, nudge_y = 0.001, size = 3, color = "black",
   #             max.overlaps = getOption("ggrepel.max.overlaps", default = 15)) +
   theme_degu() +
   theme(legend.position = "right")
plot_oni_hl

```

For cv we don't have daily values just monthly, so we have to use it by year

```{r temp_cv}

annual_study <- annual %>% 
  filter(year %in% 2009:2020)

annual_study$year <- as.factor(annual_study$year)


plot_tcv_hl <-  annual_study %>%
  ggplot( aes(x=year, y=temp_cv, group=as.factor(year), colour=as.factor(year), 
              fill=as.factor(year))) +
  geom_point( shape = 21, color="black",  size=4) +

    scale_fill_manual(values=colorsXyear) +

  scale_color_manual(values=colorsXyear) +
  gghighlight(year=="2019", #label_key = year,
              # preserve colour and modify alpha instead
              unhighlighted_params = list(colour = NULL,  size=2))+
  scale_x_continuous(limits=c(2009,2020), breaks=c(2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020) ) +
  
  labs(x="Year",  y ="T CV", colour="year") +
  
  geom_hline(yintercept = mean(annual$temp_cv), linetype="dotted") +
  geom_vline(xintercept = 5.9,  linetype="dashed", colour="gray") + #colour="lightgray",
  geom_vline(xintercept = 9.9, linetype="dashed", colour="gray") + #colour="lightgray")

   theme_degu() +
   theme(legend.position = "right")

plot_tcv_hl






```

```{r rain_cv}


range(annual$rain_cv)
plot_raincv_year <-   ggplot(annual_study, aes(x=year, y=rain_cv, group=year, 
              colour=ifelse(year == '2019', '2019', 'Other') ,
                fill=ifelse(year == '2019', '2019', 'Other'))) +
  
  geom_point( shape = 21, color="black",  size=2) +
  scale_color_manual(values = c('2019' = "#4B692DFF", 'Other' = alpha("#E1E1D2FF", 1))) +
  scale_fill_manual(values = c('2019' = "#4B692DFF", 'Other' = alpha("#E1E1D2FF", 1))) +
  #scale_y_continuous(limits=c(0, 2)) +
   geom_line( color="black") +  
  labs(x="Year", y="CV Rainfall")+
  theme_degu() +
   theme(legend.position = "none") +
  geom_hline(yintercept = mean(annual$rain_cv), linetype="dotted") 
  
plot_raincv_year

plot_raincv_hl <-  annual_study %>%
  ggplot( aes(x=year, y=rain_cv, group=as.factor(year), colour=as.factor(year), 
              fill=as.factor(year))) +
  geom_point( shape = 21, color="black",  size=4) +

    scale_fill_manual(values=colorsXyear) +

  scale_color_manual(values=colorsXyear) +
  gghighlight(year=="2019", #label_key = year,
              # preserve colour and modify alpha instead
              unhighlighted_params = list(colour = NULL,  size=2))+
  scale_x_continuous(limits=c(2009,2020), breaks=c(2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020) ) +
  
  labs(x="Year",  y ="P CV", colour="year") +
  
  geom_hline(yintercept = mean(annual$rain_cv), linetype="dotted") +
  geom_vline(xintercept = 5.9,  linetype="dashed", colour="gray") + #colour="lightgray",
  geom_vline(xintercept = 9.9, linetype="dashed", colour="gray") + #colour="lightgray")

   theme_degu() +
   theme(legend.position = "right")
 plot_raincv_hl
```


### combine plots
```{r}
ggarrange(plot_gpp_hl, plot_tcv_hl,
          plot_rain_hl,plot_raincv_hl,
          
          plot_oni_hl,
          
          nrow=3, ncol=2, align = "hv",
          common.legend = TRUE, legend = "right")

ggsave( "results/environmental covars monthXyear_v3.pdf",  width= 250, height=200, units="mm", dpi=300)

```





## plotsX season


```{r}
## temp cv 
plot_temp_sXy <- ggplot(covars_real , aes(x = season, y =temp_cv, group = season))+
  geom_col(aes(color=season, fill=season))+
  #geom_ribbon(aes(ymin = lcl, ymax = ucl, color=season2, fill=season2),  alpha = 0.3)+
  labs(x = "Season", y="CV temperature")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(#axis.text.x = element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_temp_sXy

plot_rainsum <- ggplot(covars_real , aes(x = season, y =rain_sum, group = season))+
  geom_col(aes(color=season, fill=season))+
  #geom_ribbon(aes(ymin = lcl, ymax = ucl, color=season2, fill=season2),  alpha = 0.3)+
  labs(x = "Season", y="Rainfall (mm)")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(#axis.text.x = element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(), 
       strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_rainsum

plot_rainlag <- ggplot(covars_real , aes(x = season, y =rain_sum_lag, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="Lag in rainfall (mm)")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(axis.text.x = element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         #strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_rainlag

plot_raincv <- ggplot(covars_real , aes(x = season, y =rain_cv, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="CV rainfall")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(#axis.text.x =element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_raincv

plot_raincvlag <- ggplot(covars_real , aes(x = season, y =rain_cv_lag, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="", subtitle="lag in CV rainfall")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(#axis.text.x = element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_raincvlag

plot_gpp<- ggplot(covars_real , aes(x = season, y =gpp_avg, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="GPP (Kg C/m^2/8day)")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(#axis.text.x =element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         #strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_gpp

plot_oni<- ggplot(covars_real , aes(x = season, y =oni_avg, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="El Niño index")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(#axis.text.x = element_blank(),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_oni



```

```{r}
ggarrange(plot_gpp+rremove("xlab"), 
          #plot_raincvlag+rremove("xlab"), 
           plot_rainsum+rremove("xlab"), 
          plot_temp_sXy +rremove("xlab"), 
         plot_oni +rremove("xlab"),
          #plot_rainlag+rremove("xlab"),
          plot_raincv, 
          
          nrow=5, ncol=1, align = "hv",
          common.legend = TRUE, legend = "none")

ggsave( "results/environmental covars seasonXyear_v1.pdf", width = 180, height = 200, units = "mm", dpi=300)

```


```{r}
## factors excluded from models
plot_temp<- ggplot(covars_real , aes(x = season2, y =temp_avg, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="", subtitle="Temperature (C)")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(axis.text.x = element_blank,
         legend.position="bottom",        
         legend.title = element_blank(),
         #axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_temp

plot_tempM<- ggplot(covars_real , aes(x = season2, y =temp_max, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="", subtitle="Max temperature (C)")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text.x.bottom = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_tempM

plot_lai<- ggplot(covars_real , aes(x = season2, y =lai_avg, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="", subtitle="Leaf area index")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  #facet_grid(sex~season2)+
  facet_grid(~year)+
  theme_degu() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
         strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_lai

plot_et<- ggplot(covars_real , aes(x = season2, y =et_avg, group = season))+
  geom_col(aes(color=season, fill=season))+
  labs(x = "Season", y="", subtitle="Evapotranspiration")+
  scale_fill_manual(values=c("#40B0A6", "#E1BE6A")) +
  scale_color_manual(values=c("#40B0A6", "#E1BE6A")) +
  facet_grid(~year)+
  theme_degu() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0),
         legend.position="bottom",        
         legend.title = element_blank(),
         axis.text = element_text(size = 12),
         strip.background  = element_blank(),
        strip.text  = element_blank(),
         panel.grid.major = element_blank(),
         panel.border = element_blank(),
         #axis.ticks = element_blank(),
         panel.grid.minor.x=element_blank(),
         panel.grid.major.x=element_blank() )
plot_et

```





# Climatic variation
after discussion with Praveena

https://explorador.cr2.cl/
https://climatologia.meteochile.gob.cl/application/anual/temperaturaMaximaAnual/330021/2025


plot 2000 to 2025 as we only have environmental data for this time period
use 1970 - 1999 as normal for climatic data based on daily data
use 2000 - 2008 as normal for ecosystem data




```{r}
breeding <- c(6, 7, 8, 9) # june - sep
nonbreeding <- c(10, 11, 12, 1, 2, 3, 4,5) # oct - may

period_clim <- c(1970:1999)
period_before <- c(2000:2008)
period_study <- c(2009:2020)
period_after <- c(2021:2025)

  c("1970-1999" = "#440154FF", 
    "2000-2008" = "#3B528BFF", 
    "2009-2020" = "#21908CFF", 
    "2021-2025" = "#5DC863FF", 
    "2019" = "#FDE725FF")

clim_period_colors <-  c("1970-1999" = "#CC79A7", 
    "2000-2008" = "#E69F00", 
    "2009-2020" = "#56B4E9", 
    "2021-2025" = "#009E73", 
    "2019" = "#FDE725FF")

period_colors <- c("#E69F00", "#56B4E9", "#009E73", "#FDE725FF", "#CC79A7")

```

## Temperature

```{r}
temp <- read.csv("Revision/330021_DiarioTsn_new.csv") %>% 
  separate(DateObs, c("day", "month", "year"), sep = "/", remove = FALSE) %>% 
  filter(year!="2026")

range(temp$year)
#temp$month<-str_pad(temp$month, width = 2, pad = "0")

temp$day <- as.integer(temp$day)
temp$month <- as.integer(temp$month)
temp$year <- as.integer(temp$year)

glimpse(temp)
```


### mean temperature



```{r}

t_monthly <- temp %>% 
  group_by(year, month) %>% 
  summarize(t=mean(ClimaticMean, na.rm = T))

t_annual <- temp %>% 
  group_by(year) %>% 
  summarize(t=mean(ClimaticMean, na.rm = T))%>% 
  mutate(season="annual")

t_normal <- temp %>% 
  filter(between(year, 1970, 1999)) %>% 
  summarize(t_normal=mean(ClimaticMean, na.rm = T))

tXseason <- temp %>% 
  mutate(season = if_else(month %in% breeding, "breeding", "nonbreeding")) %>% 
  group_by(year, season) %>% 
  summarize(t=mean(ClimaticMean, na.rm = T))

```

period means
```{r}
# by period
t_period <- temp %>%  
  mutate(period = case_when(
      year %in% period_clim ~ "normal",
      year %in% period_before ~ "before",
      year %in% period_study  ~ "during",
      year %in% period_after  ~ "after",
      TRUE                   ~ NA_character_
    )) %>% 
  #filter(year>1974) %>% 
  dplyr::group_by( period) %>%
  dplyr::summarize(t_avg=mean(ClimaticMean, na.rm = TRUE),
                   t_sd=sd(ClimaticMean, na.rm = TRUE),
                   n_years = n(),
                   t_se = t_sd / sqrt(n_years), .groups = "drop") %>% 
  add_row(period = "2019", 
          t_avg = 15.97068,
          t_sd = NA_real_,    # or specific values
          n_years = 1,
          t_se = NA_real_) %>% 
  filter(period!="NA")

t_period$period<- factor(t_period$period, 
                         levels=c("normal", "before", "during", "after", "2019"),
                         labels=c("1970-1999", "2000-2008", "2009-2020", "2021-2025", "2019"))

plot_t_period <- ggplot(t_period, aes(x = period, y = t_avg, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = t_avg - t_se, ymax = t_avg + t_se),
                width = 0.2, linewidth = 0.8) +

  #scale_fill_manual(values = c("before" = "#7D593D", "during" = "lightgrey", "after" = "#AF9153")) +
  #scale_fill_viridis(discrete = TRUE, option = "cividis") +
  scale_fill_manual(values = clim_period_colors) + 
  #scale_fill_viridis(discrete = TRUE, option = "D", name = "Period")+

  labs(x = "Period", y = " ", subtitle = "Period means ± SE" ) +#, y = "mean T"~(degree*C)       title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "none",
        #axis.text.x = element_text(angle = 45, vjust = 0.5),
        #axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = margin(2,2,2,2, "pt"))

plot_t_period 
```


### max temperature

```{r max}

tmax_monthly <- temp %>% 
  group_by(year, month) %>% 
  summarize(tmax=max(MaxTemperature, na.rm = T))

tmax_annual <- temp %>% 
  group_by(year) %>% 
  summarize(tmax=max(MaxTemperature, na.rm = T))%>% 
  mutate(season="annual")

tmax_normal <- temp %>% 
  filter(between(year, 1970, 1999)) %>% 
  summarize(tmax_normal=max(MaxTemperature, na.rm = T)) %>% 
  pull(tmax_normal)

tmaxXseason <- temp %>% 
  mutate(season = if_else(month %in% breeding, "breeding", "nonbreeding")) %>% 
  group_by(year, season) %>% 
  summarize(tmax=max(MaxTemperature, na.rm = T))


maxtemps <- temp %>% 
  filter(year>2008, 
         MaxTemperature>37) 

```

```{r mean}

# tmax_monthly <- temp %>% 
#   group_by(year, month) %>% 
#   summarize(tmax=mean(MaxTemperature, na.rm = T))
# 
# tmax_annual <- temp %>% 
#   group_by(year) %>% 
#   summarize(tmax=mean(MaxTemperature, na.rm = T))%>% 
#   mutate(season="annual")
# 
# tmax_normal <- temp %>% 
#   filter(between(year, 1970, 1999)) %>% 
#   summarize(tmax_normal=mean(MaxTemperature, na.rm = T))
# 
# tmaxXseason <- temp %>% 
#   mutate(season = if_else(month %in% breeding, "breeding", "nonbreeding")) %>% 
#   group_by(year, season) %>% 
#   summarize(tmax=mean(MaxTemperature, na.rm = T))

```


period max
```{r}
# by period
tmax_period <- temp %>% 
   mutate(period = case_when(
      year %in% period_clim ~ "normal",
      year %in% period_before ~ "before",
      year %in% period_study  ~ "during",
      year %in% period_after  ~ "after",
      TRUE                   ~ NA_character_
    )) %>% 
  #summarize(tmax = max(MaxTemperature, na.rm = T)) %>% 
   dplyr::group_by( period) %>%
  dplyr::summarize(tmax_avg=max(MaxTemperature, na.rm = T),
                   tmax_sd=sd(tmax_avg, na.rm = TRUE),
                   n_years = n(),
                   tmax_se = tmax_sd / sqrt(n_years), .groups = "drop") %>% 
  add_row(period = "2019", 
          tmax_avg = 39.3,
          tmax_sd = NA_real_,    # or specific values
          n_years = 1,
          tmax_se = NA_real_)

tmax_period$period<- factor(tmax_period$period, 
                         levels=c("normal", "before", "during", "after", "2019"),
                         labels=c("1970-1999", "2000-2008", "2009-2020", "2021-2025", "2019"))

plot_tmax_period <- ggplot(tmax_period, aes(x = period, y = tmax_avg, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = tmax_avg - tmax_se, ymax = tmax_avg + tmax_se),
                width = 0.2, linewidth = 0.8) +

  scale_fill_manual(values = clim_period_colors) +  
  
  labs(x = "Period", y = " " ) +#,  y = "max T"~(degree*C)      title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "none",
        #axis.text.x = element_text(angle = 45, vjust = 0.5),
        #axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = margin(2,2,2,2, "pt"))

plot_tmax_period 
```

### annual

```{r}

range(t_annual$t)

Tannual <- t_annual %>% 
  left_join(tmax_annual) %>% 
  filter(year>1999) %>% 
  select(-season) %>% 
  pivot_longer(cols = c(t, tmax), 
               names_to = "type", 
               values_to = "temp")

head(Tannual)

Tannual$type <- factor(Tannual$type, 
                         levels=c("t", "tmax"),
                         labels=c("mean T", "max T"))

scale_factor <- with(Tannual, 
                     max(temp[type == "mean T"], na.rm = TRUE) /
                     max(temp[type == "max T"],  na.rm = TRUE))


plotTyear_axes <- ggplot(Tannual, aes(x = year)) +
  # mean T (primary axis)
  geom_line(data = subset(Tannual, type == "mean T"),
            aes(y = temp, color = "mean T")) +
  geom_point(data = subset(Tannual, type == "mean T"),
             aes(y = temp, color = "mean T")) +
  # max T (secondary axis, scaled to primary)
  geom_line(data = subset(Tannual, type == "max T"),
            aes(y = temp * scale_factor, color = "max T")) +
  geom_point(data = subset(Tannual, type == "max T"),
             aes(y = temp * scale_factor, color = "max T")) +
   # dashed normals
  geom_hline(yintercept = t_normal$t,
             linetype = "dashed", color = "#fd8d3c") +  # mean T normal (primary axis)

  geom_hline(yintercept = tmax_normal * scale_factor,
             linetype = "dashed", color = "#882255") +  # max T normal (secondary axis; scaled) 
  # scale_y_continuous(
  #   name = expression("Mean T ("*degree*C*")"),
  #   limits = c(13.5, 16), 
  #   sec.axis = sec_axis(
  #     trans = ~ . / scale_factor,
  #     #name  = expression("Max T ("*degree*C*")")
  #   )
    
    scale_y_continuous(
  name = expression("Mean T ("*degree*C*")"),
  limits = c(13.5, 16),  # Keep primary as-is if it fits data; adjust if needed
  expand = expansion(mult = c(0, 0)),  # Remove padding for tight fit
  sec.axis = sec_axis(
    trans = ~ . / scale_factor,
    #name = expression("Max T ("*degree*C*")"),  # Uncomment/fix this
    breaks = c(34,35, 36,37, 38,39, 40)  # Custom ticks at desired range
  )
) +
coord_cartesian(ylim = c(13.5, 16)) + # Ensures view clips to these limits
   
  scale_color_manual(values = c("mean T" = "#fd8d3c", "max T" = "#882255")) +
  theme_classic() +
  labs(x = "Year", subtitle = "Annual temperatures" ) +
  scale_x_continuous(breaks = seq(2000, 2025, 5),
                     expand = expansion(add = c(0.5, 0.5))) +
  geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
    theme( # Right y-axis line & ticks: match anomaly line color
    axis.line.y.right = element_line(color = "#882255"),
    axis.ticks.y.right = element_line(color = "#882255"),
    # Right y-axis text: match line color
    axis.text.y.right = element_text(color = "#882255"),
    axis.title.y.right = element_text(color = "#882255"), 
    legend.position = "none")
plotTyear_axes

ggsave("./Revision/Tannual.pdf", width=13.5, height=5, units = "cm", dpi=600)

```


```{r}
plotT_legend <- ggplot(Tannual, aes(x = year, y = temp, group = type, color = type)) +
  geom_line(linetype="solid") + 
  geom_point() +
  labs(x = "Year", 
       y = expression("Annual temperature"~(degree*C)), 
       subtitle = "Annual and seasonal variation") +
  # Hlines WITH aes() for legend inclusion
  geom_hline(aes(yintercept = tmax_normal$tmax, color = "Max T threshold", linetype = "Baseline mean"), 
             linewidth = 0.8) +
  geom_hline(aes(yintercept = t_normal$t, color = "Mean T threshold", linetype = "Baseline mean"), 
             linewidth = 0.8) +
 geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
  # Combined scales
  scale_color_manual(
    values = c("mean T" = "#fd8d3c", "max T" = "#882255",           # Your 2 types
               "Mean T threshold" = "#fd8d3c",  # Match line colors
               "Max T threshold" = "#882255"),
    name = NULL  # No legend title
  ) +
  scale_linetype_manual(values = c("solid" = "solid",
    "Baseline mean" = "dashed"), name = NULL) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  theme_classic() +
  theme(legend.title = element_blank())

plotT <- ggplot(Tannual, aes(x = year, y = temp, group = type, color = type)) +
  geom_line(linetype="solid") + 
  geom_point() +
  labs(x = "Year", 
       y = expression("Annual temperature"~(degree*C)), 
       subtitle = "Annual and seasonal variation") +
  # Hlines WITH aes() for legend inclusion
  geom_hline(aes(yintercept = tmax_normal$tmax, color = "Max T threshold", linetype = "Baseline mean"), 
             linewidth = 0.8) +
  geom_hline(aes(yintercept = t_normal$t, color = "Mean T threshold", linetype = "Baseline mean"), 
             linewidth = 0.8) +
 geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
  # Combined scales
  scale_color_manual(
    values = c("mean T" = "#fd8d3c", "max T" = "#882255",           # Your 2 types
               "Mean T threshold" = "#fd8d3c",  # Match line colors
               "Max T threshold" = "#882255"),
    name = NULL  # No legend title
  ) +
  scale_linetype_manual(values = c("solid" = "solid",
    "Baseline mean" = "dashed"), name = NULL) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  theme_classic() +
  theme(legend.title = element_blank(), 
        legend.position = "none")

```


### seasonal

```{r}
Tseason <- tXseason %>% 
  left_join(tmaxXseason) %>% 
  filter(year>1999) %>% 
   pivot_longer(cols = c(t, tmax), 
               names_to = "type", 
               values_to = "temp") %>% 
  #unite(group, season, type, remove = FALSE) %>% 
   mutate(
    season_type = interaction(season, type),  # Full grouping factor
    group = case_when(  # FIXED: Create unique labels directly
      season == "breeding" & type == "t" ~ "Breeding mean",
      season == "breeding" & type == "tmax" ~ "Breeding max", 
      season == "nonbreeding" & type == "t" ~ "Nonbreed mean",
      TRUE ~ "Nonbreed max"
    )

  ) %>%
  arrange(year, season_type)

head(Tseason)
range(Tseason$temp[Tseason$type=="t"])
range(Tseason$temp[Tseason$type=="tmax"])

print(table(Tseason$group, useNA = "always"))  # Should show 4 levels with counts

```

```{r}
# Scale factor: map max T range to mean T range
scale_factor <- with(Tseason, 
                     max(temp[group %in% c("Breeding mean", "Nonbreed mean")], na.rm = TRUE) /
                     max(temp[group %in% c("Breeding max", "Nonbreed max")], na.rm = TRUE))


```


```{r}
# Your desired right‑axis labels (max T)
right_labels <- c(25, 30, 35, 40)
# Corresponding left‑axis values (mean T)
left_vals    <- 0.7333 * right_labels - 10.33  # approx: 8, 11.7, 15.3, 19

# Left‑axis range (your data)
mean_temp_range <- range(Tseason$temp[Tseason$type == "t"], na.rm = TRUE)
# should be about c(7.85, 18.83)

# For plotting, keep max lines scaled to place 25–40 on the right axis
scaling_for_plot <- 0.7333  # or 11/15
offset_for_plot  <- -10.33

# desired mapping: left ≈ a*right + b
a <- 11/15  # ≈0.733
b <- 8 - a*25

# corresponding inverse for sec_axis
sec_trans <- function(x) (x - b)/a   #  → right = (left - b)/a


# Max data must be plotted on the same scale as the left axis
# so they appear at ≈8–19 °C (mean‑like scale) even though they come from max T
Tseason_max_scaled <- Tseason %>%
  filter(grepl("max", group)) %>%
  mutate(y_plot = temp * a + b)   # project max T to desired left‑axis positions

plotTseason_axes <- ggplot(Tseason, aes(x = year, group = season_type)) +
  # Mean lines (primary axis, unchanged)
  geom_line(data = filter(Tseason, grepl("mean", group)),
            aes(y = temp, color = group)) + 
  geom_point(data = filter(Tseason, grepl("mean", group)),
             aes(y = temp, color = group, shape = season)) +
  # Max lines: now mapped to 8–19 °C on left axis
  geom_line(data = Tseason_max_scaled,
            aes(y = y_plot, color = group)) + 
  geom_point(data = Tseason_max_scaled,
             aes(y = y_plot, color = group, shape = season)) +
  labs(x = "Year", y = expression("Mean T"~(degree*C)), 
       subtitle = "Seasonal temperatures") +
  scale_color_manual(values = c("Breeding mean" = "#fd8d3c42", 
                                "Breeding max"  = "#88225542",
                                "Nonbreed mean" = "#fd8d3c", 
                                "Nonbreed max"  = "#882255")) +
  scale_shape_manual(values = c("breeding" = 19,      # ● circle
                                "nonbreeding" = 15), # ■ square
                     name = "Season") +
  scale_y_continuous(
    limits   = c(7, 20),   # just enough to cover 8–19
    breaks   = pretty(c(7, 20), n = 6),
    sec.axis = sec_axis(
      ~ sec_trans(.),      # or ~ (. - b)/a
      name   = expression("Max T"~(degree*C)),
      breaks = right_labels
    )
  ) +
  theme_classic() +
  scale_x_continuous(breaks = seq(2000, 2025, 5),
                     expand = expansion(add = c(0.5, 0.5))) +
  geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), 
             alpha = 0.5) + 
  theme(
    axis.line.y.right  = element_line(color = "#882255"),
    axis.ticks.y.right  = element_line(color = "#882255"),
    axis.text.y.right   = element_text(color = "#882255"),
    axis.title.y.right  = element_text(color = "#882255"),
    legend.position     = "bottom",
    plot.margin         = margin(2, 2, 2, 2, "pt")
  )


plotTseason_axes 
```



```{r}
plotTseason_legend <- ggplot(Tseason,  aes(x=year, y=temp,  color=group, group=season_type)) +
  geom_line() + geom_point()+
  labs(x = "Year", y="Seasonal T"~(degree*C), subtitle="")+
  #scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +
  scale_color_manual(values = c( "#fd8d3c42", "#88225542", "#fd8d3c", "#882255")) + #  "#EF674E" 
  #scale_color_viridis_d(option = "plasma", n=4)+
  theme_classic() +
  scale_x_continuous(breaks = seq(2000, 2025, 5), 
                     expand = expansion(add = c(0.1, 0.1))) +
  #geom_hline(yintercept = tmax_normal$tmax, linetype="dashed", color="#882255") +
  #geom_hline(yintercept = t_normal$t, linetype="dashed", color="#fd8d3c") +
  geom_vline(xintercept = 2019, color = alpha("red", 0.5)) +
  geom_vline(xintercept = c(2008.5, 2019.5), linetype = "dotted", color = alpha("black", 0.5)) +
  theme(legend.title = element_blank(), 
        legend.position = "right")


plotTseason <- ggplot(Tseason,  aes(x=year, y=temp,  color=group, group=season_type)) +
  geom_line() + geom_point()+
  labs(x = "Year", y="Seasonal T"~(degree*C), subtitle="")+
  #scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +
  scale_color_manual(values = c( "#fd8d3c42", "#88225542", "#fd8d3c", "#882255")) + #  "#EF674E" 
  #scale_color_viridis_d(option = "plasma", n=4)+
  theme_classic() +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  #geom_hline(yintercept = tmax_normal$tmax, linetype="dashed", color="#882255") +
  #geom_hline(yintercept = t_normal$t, linetype="dashed", color="#fd8d3c") +
 geom_vline(xintercept = c(2008.5, 2019, 2019.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
  theme(legend.title = element_blank(), 
        legend.position = "none")
plotTseason
```




## Precipitation sum

add line for cumulative anomaly (mm) as secondary axis

The cumulative value  as anomaly = annual value − baseline mean.

```{r}
# p_daily <- read.csv("Revision/psum_daily.csv") %>% 
#   rename(year=agno, 
#          month=mes,
#          day=dia, 
#          psum=valor)

p_daily <- read.csv("Revision/330021_DiarioRR.csv") %>% 
  separate(Fecha, c("day", "month", "year"), sep = "-", remove = FALSE) %>% 
  rename(p=Total)

p_daily$day <- as.integer(p_daily$day)
p_daily$month <- as.integer(p_daily$month)
p_daily$year <- as.integer(p_daily$year)
glimpse(p_daily)

p_monthly <- p_daily %>% 
  group_by(year, month) %>% 
  summarize(p=sum(p))

p_annual <- p_daily %>% 
  group_by(year) %>% 
  summarize(p=sum(p)) %>% 
  mutate(season="annual")

p_normal <- p_annual %>% 
  filter(between(year, 1970, 1999)) %>% 
  summarize(pmean=mean(p, na.rm=T)) %>% 
  pull(pmean)


pXseason <- p_monthly %>% 
  mutate(season = if_else(month %in% breeding, "breeding", "nonbreeding")) %>% 
  group_by(year, season) %>% 
  summarize(p=sum(p))

p_anom <- p_annual %>% 
  filter(between(year, 2000, 2025)) %>% 
  mutate(season="annual", 
         pct_anomaly = 100 * (p - p_normal) / p_normal,
         anomaly=p-p_normal,
         cum_p_deficit = cumsum(anomaly)) #cumsum(pmax(0, -p_anom))

# scale factor 
#scale_factor <- max(pXseason$p, na.rm = TRUE) / max(p_anom$cum_p_deficit, na.rm = TRUE)
#scale_factor <- range(pXseason$p, na.rm = TRUE)[2] / range(p_anom$cum_p_deficit, na.rm = TRUE)[2]
# Then use in sec_axis(~ . / scale_factor, ...) or multiply anomaly y-values

# Define exact ranges
precip_max <- 500
anomaly_range <- c(-2000, 500)

# Scale factor maps anomaly range to precip range
scale_factorp <- precip_max / diff(anomaly_range)
offset <- 0 - anomaly_range[1] * scale_factorp  # Left 0 -> anomaly 0? No, independent

range(p_anom$cum_p_deficit)

p_anom %>% 
  filter(year == 2019) %>%
  mutate( statement = sprintf("precipitation was %.1f%% %s than the 1975-1999 mean",
                             abs(pct_anomaly), ifelse(pct_anomaly < 0, "lower", "higher")))   
p_anom %>% 
  filter(between(year, 2009, 2020)) %>%
  summarize(mean_anomaly = mean(pct_anomaly))   



```
```{r}



plotp <- pXseason %>% 
  bind_rows(p_annual) %>% 
  filter(between(year,2000,2025)) %>% 
  
  ggplot( aes(x=year, y=p, fill = season)) +
  geom_col(position = position_dodge2(width = 0.8, padding = 0.05)) + 
  scale_x_continuous(breaks = seq(2000, 2025, 5), 
                     expand = expansion(add = c(0.0, 0.0))) +
  
  # Cumulative line from DIFFERENT data frame
  geom_line(data = p_anom, 
            mapping = aes(x = year, 
                          y = (cum_p_deficit- anomaly_range[1]) * scale_factorp), 
            color = "#1E88E5",                    
            linewidth = 1, 
            inherit.aes = FALSE) +                 # Ignore main plot aes
  
  labs(x = "Year", y = "P (mm)", subtitle = "") +

  geom_hline(yintercept = p_normal, 
             linetype = "dashed", color="#5A5C8C") +
  geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
  scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +

  scale_y_continuous(
    name = expression("P (mm)"),
    limits = c(0, 500),                   # LEFT: Fixed 0-500
    expand = c(0, 0),
    sec.axis = sec_axis(transform = ~ (. / scale_factorp) + anomaly_range[1],  # Reverse to true anomaly
                        #name = "Cumulative Anomaly (mm)",
                        breaks = seq(-2000, 500, 500) )
  ) +   
  theme_classic() +
  theme( # Right y-axis line & ticks: match anomaly line color
    axis.line.y.right = element_line(color = "#1E88E5"),
    axis.ticks.y.right = element_line(color = "#1E88E5"),
    # Right y-axis text: match line color
    axis.text.y.right = element_text(color = "#1E88E5"),
    axis.title.y.right = element_text(color = "#1E88E5"), 
    legend.position = "none",
        plot.margin = margin(2,2,2,2, "pt")) 
plotp




```

```{r}
plotp_legend <- pXseason %>% 
  bind_rows(p_annual) %>% 
  filter(year > 1999) %>% 
  ggplot(aes(x = year, y = p, group = season)) +
  geom_col(aes(color = season, fill = season), position = position_dodge(width = 0.8)) +
  # Cumulative line WITH aes() for legend
  geom_line(data = p_anom, 
            aes(x = year, 
                y = (cum_p_deficit - anomaly_range[1]) * scale_factor, 
                color = "Cumulative P anomaly",
                linetype = "Cumulative anomaly"),  # Legend factors
            linewidth = 1, 
            inherit.aes = FALSE) +
  labs(x = "Year", y = "P (mm)", subtitle = "") +
  geom_hline(aes(yintercept = p_normal, color = "Normal P", linetype = "1970-1999 mean"),
             linewidth = 0.8) +
  geom_vline(xintercept = c(2008.5, 2019, 2019.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
  
  # Extended scales
  scale_fill_manual(values = c("#5A5C8C", "#40B0A6", "#E1BE6A")) +
  scale_color_manual(
    values = c("#5A5C8C", "#40B0A6", "#E1BE6A",            # Your seasons
               "Cumulative P anomaly" = "#1E88E5",           # Blue line
               "Normal P" = "gray50"),                       # Hline color
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Cumulative anomaly" = "solid", "1970-1999 mean" = "dashed"),
    name = NULL
  ) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  scale_y_continuous(
    name = expression("P (mm)"),
    limits = c(0, 500),
    expand = c(0, 0),
    sec.axis = sec_axis(~ (. / scale_factor) + anomaly_range[1],
                        name = "Cumulative Anomaly (mm)",
                        breaks = seq(-2000, 500, 500))
  ) +   
  theme_classic() +
  theme(
    axis.line.y.right = element_line(color = "#1E88E5"),
    axis.ticks.y.right = element_line(color = "#1E88E5"),
    axis.text.y.right = element_text(color = "#1E88E5"),
    axis.title.y.right = element_text(color = "#1E88E5"), 
    legend.position = "right",
        plot.margin = margin(2,2,2,2, "pt")
  )
plotp_legend
```
```{r}
plotp_simple <- pXseason %>% 
  bind_rows(p_annual) %>% 
  filter(year>1999) %>% 
  
  ggplot( aes(x=year, y=p, fill = season)) +
  geom_col(position = position_dodge2(width = 0.8, padding = 0.05)) + 
  scale_x_continuous(breaks = seq(2000, 2025, 5), 
                     expand = expansion(add = c(0.1, 0.1))) +
  
  labs(x = "Year", y = "P (mm)", subtitle = "") +

  geom_hline(yintercept = p_normal, 
             linetype = "dashed", color="#5A5C8C") +
  geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5) +
  scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +

  
  theme_classic() +
  theme(  
    legend.position = "none",
        plot.margin = margin(2,2,2,2, "pt")) 
plotp_simple
```


period means
```{r}
# by period
p_period <- p_annual%>%  
  filter(between(year, 1970, 2025)) %>% 
  mutate(period = case_when(
      year %in% period_clim ~ "normal",
      year %in% period_before ~ "before",
      year %in% period_study  ~ "during",
      year %in% period_after  ~ "after",
      TRUE                   ~ NA_character_
    )) %>% 
  
  dplyr::group_by( period) %>%
  dplyr::summarize(p_sum=sum(p, na.rm = TRUE),
                   pmean=mean(p, na.rm = TRUE),
                   p_sd=sd(p, na.rm = TRUE),
                   n_years = n(),
                   p_se = p_sd / sqrt(n_years), .groups = "drop") %>% 
  add_row(period = "2019", 
          pmean = 47.2,
          p_sd = NA_real_,  
          n_years = 1,
          p_se = NA_real_)

p_period$period<- factor(p_period$period, 
                         levels=c("normal", "before", "during", "after", "2019"),
                         labels=c("1970-1999", "2000-2008", "2009-2020", "2021-2025", "2019"))


plot_p_period <- ggplot(p_period, aes(x = period, y = pmean, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = pmean - p_se, ymax = pmean + p_se),
                width = 0.2, linewidth = 0.8) +

scale_fill_manual(values = clim_period_colors) + 
  
  labs(x = "Period", y="" ) +#,    , y = "P"    title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "none", 
        legend.direction = "vertical",
        legend.title = element_blank(),  # No "Period" title
        legend.text = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1),
        plot.margin = margin(2,2,2,2, "pt"))


plot_p_period 


```

```{r}
plot_p_period_legend <- ggplot(p_period, aes(x = period, y = p_sum, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = p_sum - p_se, ymax = p_sum + p_se),
                width = 0.2, linewidth = 0.8) +

scale_fill_manual(values = clim_period_colors) + 
  
  labs(x = "Period", y="" ) +#,    , y = "P"    title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "bottom", 
        legend.direction = "vertical",
        legend.title = element_blank(),  # No "Period" title
        legend.text = element_text(size = 10),
        axis.title.x = element_blank())
        #axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
```

## ENSO
```{r}

oni <- read.csv("Revision/oni.csv") %>% 
  separate(Date, c("year", "month", "day"), remove =FALSE) %>% 
  unite("ym", year:month, remove = FALSE) %>% 
  select(-day)

colnames(oni)[5] <- "oni"

oni$year<-as.integer(oni$year)
oni$month<-as.integer(oni$month)

oni <- oni %>% 
  filter(between(year, 2000, 2025)) %>% 
  mutate(date = ymd(ym, truncated = 2),
         Date=as.Date(Date)) %>% 
  ungroup() %>% 
  arrange(date)  
glimpse(oni)
head(oni)
range(oni$Date)

plot_oni_year <- oni %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(oni_avg=mean(oni)) %>% 
  ggplot( aes(x=year, y=oni_avg)) +
  geom_line(color="#bdbdbd") +  
  geom_point( shape = 21, color="black", fill="#bdbdbd", size=2 ) +
  theme_classic() +
  labs(x="Year", y= "El Niño index") +
  #geom_hline(yintercept = mean(oni$oni, na.rm=T), linetype="dashed") +
  scale_x_continuous(limits=c(2000, 2025), 
                     breaks = seq(2000, 2025, 5),
                     expand = expansion(add = c(0.5, 0.5))) +
  geom_vline(xintercept = 2019,  color=alpha("red", 0.5)) + #linetype = "dashed",
  geom_vline(xintercept = 2008.5, linetype = "dotted", color=alpha("black", 0.5)) +
  geom_vline(xintercept = 2020.5, linetype = "dotted", color=alpha("black", 0.5)) 
  
plot_oni_year



plot_oni <- ggplot(oni, aes(x = date, y = oni)) +
  geom_line() +
  labs(x = "Year", y = "Monthly ONI") +
  theme_classic() +
  scale_x_date(
    limits  = c(as.Date("2000-01-01"), as.Date("2025-12-01")),
    breaks  = seq(as.Date("2000-06-15"), as.Date("2025-6-15"), by = "5 years"),
    labels  = c("2000", "2005", "2010", "2015", "2020", "2025"),
    expand  = c(0.01, 0.01)   # small padding, not the 0.5 stretch you had
  ) +
  # align vertical lines with the numeric‑year plots (2008.5, 2019, 2020.5)
  geom_vline(xintercept = as.Date("2008-12-01"),   # close to 2008.5
             linetype = "dotted",
             color = alpha("black", 0.5)) +
  geom_vline(xintercept = as.Date("2019-07-01"),   # 2019 mid‑year
             color = alpha("red", 0.5)) +
  geom_vline(xintercept = as.Date("2020-12-31"),   # close to 2020.5
             linetype = "dotted",
             color = alpha("black", 0.5)) +
  theme(
    plot.margin = margin(2, 2, 2, 2, "pt")
  )
plot_oni 

ggarrange(plot_oni, plot_oni_year, ncol=1, align = "v")
```


```{r}
df <- oni %>%
  mutate(date = ymd(ym, truncated = 2)) %>%
  arrange(date) %>%
  mutate(
    oni_3m = zoo::rollmean(oni, k = 3, fill = NA, align = "right")
  ) %>%
  mutate(enso_flag = case_when(oni_3m >= 0.5 ~ 1,
                               oni_3m <= -0.5 ~ -1,
                               TRUE ~ 0),
    # Identify runs of 5+ consecutive qualifying periods
    event_run = rle(enso_flag)$lengths,
    event_id = cumsum(enso_flag != lag(enso_flag, default = 0)),
    # Only keep events with 5+ consecutive qualifying 3-month periods
    is_event = ave(event_run, event_id, FUN = function(x) x >= 5),
    enso_event = case_when(
      enso_flag == 1 & is_event ~ "El Niño",
      enso_flag == -1 & is_event ~ "La Niña",
      TRUE ~ NA_character_
    )
  )


df_clean <- oni %>%
  mutate(date = ymd(ym, truncated = 2)) %>%
  arrange(date) %>%
  filter(!is.na(oni), !is.na(date))  # Remove any NAs first

# 3-month ONI (safe length)
df_clean$oni_3m <- zoo::rollmean(df_clean$oni, k = 3, fill = NA, align = "right")

# Safe flags (replace NA explicitly)
df_clean$enso_flag <- ifelse(is.na(df_clean$oni_3m), 0,
                           ifelse(df_clean$oni_3m >= 0.5, 1, 
                                  ifelse(df_clean$oni_3m <= -0.5, -1, 0)))

# Run length encoding SAFELY (no NAs)
df_clean$run_id <- cumsum(c(TRUE, df_clean$enso_flag[-1] != df_clean$enso_flag[-length(df_clean$enso_flag)]))
df_clean$event_periods <- ave(df_clean$enso_flag, df_clean$run_id, FUN = function(x) length(x[x != 0]))

# FIXED subsetting (remove NAs first)
valid_rows <- !is.na(df_clean$event_periods) & df_clean$enso_flag != 0 & df_clean$event_periods >= 5
df_clean$enso_event <- NA_character_
df_clean$enso_event[valid_rows] <- ifelse(df_clean$enso_flag[valid_rows] > 0, "El Niño", "La Niña")

# Verify lengths match
print(nrow(df_clean))
print(length(df_clean$enso_event))

ggplot(df_clean, aes(x = date, y = oni)) +
  geom_line() +
  #geom_point() +
  
  geom_point(aes(color = enso_event), size = 1) +
  scale_color_manual(values = c("El Niño" = alpha("red", 0.5), "La Niña" = alpha("blue", 0.5), "Neutral" = "gray"))+
  
  # geom_rect(data = filter(df_clean, !is.na(enso_event)),
  #           aes(xmin = date - days(30), xmax = date + days(30),
  #               ymin = -Inf, ymax = Inf, fill = enso_event),
  #           alpha = 0.3, inherit.aes = FALSE) +
  # scale_fill_manual(values = c("El Niño" = alpha("red", 0.1), "La Niña" = alpha("blue", 0.1))) +
  
    geom_hline(yintercept = c(-0.5, 0, 0.5), linetype = "dashed", color = "gray50") +
  scale_x_date(breaks = seq(as.Date("2000-01-01"), as.Date("2025-01-01"), by = "5 years"),
               labels = c("2000", "2005", "2010", "2015", "2020", "2025")) +
  labs(x = "Year", y = "Monthly ONI") +
  theme_classic() +
  geom_vline(xintercept = as.Date("2019-01-01"),  color=alpha("red", 0.5)) + #linetype = "dashed",
  geom_vline(xintercept = as.Date("2009-01-01"), linetype = "dotted", color=alpha("black", 0.5)) +
  geom_vline(xintercept = as.Date("2021-01-01"), linetype = "dotted", color=alpha("black", 0.5)) 



```


## get plots aligned



```{r}
ggarrange(plotTyear_axes,
          plotTseason_axes,
          plotp, 
          plot_oni,
          ncol=1, align = "hv")

ggarrange(plotTyear_axes,
          plotTseason_axes,
          plotp, 
          plot_oni_year,
          ncol=1, align = "hv")
```

## combine plots
Ecology: maximum allowable size (18 cm wide by 22 cm tall) 

```{r}
climate <- ggarrange(plotTyear_axes + rremove("xlab"),    plot_t_period + rremove("xlab"),
                     plotTseason_axes  + rremove("xlab"), plot_tmax_period + rremove("xlab"),
                     plotp + rremove("xlab"), plot_p_period ,
                     plot_oni,
                     nrow=4, ncol=2, align = "hv", 
                     widths = c(5,1), #  widths = c(3,1) period widths = c(2.5,1)
                     legend = "none")
climate
ggsave("./Revision/climate_variation_v2.pdf", width=18, height=20, units = "cm", dpi=300)

ggexport(climate, filename = "./Revision/climate_variation.pdf", 
         width = 18, height = 20, units = "cm", dpi = 300)

climate_legends <- ggarrange(plotT + rremove("xlab"),    plot_t_period + rremove("xlab"),
          plotTseason  + rremove("xlab"), plot_tmax_period + rremove("xlab"),
          plotp_simple + rremove("xlab"), plot_p_period ,
          plot_oni,
          nrow=4, ncol=2, align = "hv", widths = c(2,1), #  widths = c(3,0.5) period widths = c(2,1) 
            legend = "right")
climate_legends
ggsave("./Revision/climate_variation_legends.pdf", width=18, height=20, units = "cm", dpi=600)


```



# Ecosystem variation

```{r}
# Colorblind-safe divergent palette for 5 time periods
period_colors <- c("#E69F00", "#56B4E9", "#009E73", "#FDE725FF", "#CC79A7")
```


MODIS
use 2000 - 2008 as normal for ecosystem data

```{r}

modis <- read.csv("Revision/Modis_monthly_data.csv") %>% 
   dplyr::rename(year=Year,
        month=Month) 
range(modis$year)


modis_day <- read.csv("Revision/modis_daily_ET_GPP_LAI.csv")  %>% 
  dplyr::rename(year=Year,
        month=Month) 

modis_annual <- modis_day %>% 
  group_by(year) %>% 
  summarize(gpp=mean(GPP, na.rm=T),
            et=mean(ET, na.rm=T),
            lai=mean(LAI, na.rm=T))%>% 
  mutate(season="annual") %>% 
  filter(year!="")

modisXseason <- modis %>% 
  mutate(season = if_else(month %in% breeding, "breeding", "nonbreeding")) %>% 
  group_by(year, season) %>% 
  summarize(gpp=mean(GPP_avg, na.rm=T),
            et=mean(ET_avg, na.rm=T),
            lai=mean(LAI_avg, na.rm=T))

# by period
modis_period <- modis_annual %>%  
  mutate(period = case_when(
      #year %in% period_clim ~ "longterm",
      year %in% period_before ~ "before",
      year %in% period_study  ~ "during",
      year %in% period_after  ~ "after",
      TRUE                   ~ NA_character_
    )) %>% 
  #filter(year>1974) %>% 
  dplyr::group_by( period) %>%
  dplyr::summarize(n_years = n(),
                   gpp_avg=mean(gpp, na.rm = TRUE),
                   gpp_sd=sd(gpp, na.rm = TRUE),
                   gpp_se = gpp_sd / sqrt(n_years),
                   
                   et_avg=mean(et, na.rm = TRUE),
                   et_sd=sd(et, na.rm = TRUE),
                   et_se = et_sd / sqrt(n_years),
                   
                   lai_avg=mean(lai, na.rm = TRUE),
                   lai_sd=sd(lai, na.rm = TRUE),
                   lai_se = lai_sd / sqrt(n_years),
                   
                   .groups = "drop") %>% 
  add_row(period = "2019", 
          gpp_avg = 0.9032609,
          et_avg = 0.2717391, 
          lai_avg = 0.2663043,
          n_years = 1)

modis_period$period<- factor(modis_period$period, 
                         levels=c( "before", "during", "after", "2019"),
                         labels=c("2000-2008", "2009-2020", "2021-2025", "2019"))
```


### gpp

```{r}
gpp_normal <- modis_day %>% 
  filter(between(year, 2000, 2008)) %>% 
  summarize(gppmean=mean(GPP, na.rm=T)) %>% 
  pull(gppmean)

gpp_anom <- modis_day %>% 
  filter(year>1999) %>% 
  mutate(season="annual", 
         pct_anomaly = 100 * (GPP - gpp_normal) / gpp_normal,
         anomaly=GPP-gpp_normal,
         cum_p_deficit = cumsum(anomaly)) #cumsum(pmax(0, -p_anom))

gpp_anom %>% 
  filter(year == 2019) %>%
  select(pct_anomaly, anomaly) %>% 
  mutate( statement = sprintf("GPP was %.1f%% %s than the 2000-2008 mean",
                             abs(pct_anomaly), ifelse(pct_anomaly < 0, "lower", "higher")))   
gpp_anom %>% 
  filter(between(year, 2009, 2020)) %>%
  summarize(mean_pct_anomaly = mean(pct_anomaly),
            mean_anomaly = mean(anomaly))   
```


```{r}
plot_gpp <-modisXseason %>% 
  bind_rows(modis_annual) %>% 
  filter(year>1999) %>% 
  ggplot( aes(x=year, y=gpp, fill=season)) +
  geom_col(position = position_dodge2(width = 0.8, padding = 0.05)) + 
  scale_x_continuous(breaks = seq(2000, 2025, 5), 
                     expand = expansion(add = c(0.1, 0.1))) +
  labs(x = "Year", y=expression(GPP~(gC/m^2)), subtitle="Annual and seasonal variation")+
  scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +

  theme_classic() +
  theme(legend.position = "none",
        plot.margin = margin(2,2,2,2, "pt")) +

  geom_hline(yintercept = gpp_normal, linetype="dashed", color = "#5A5C8C") +
   geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5)
plot_gpp
```

```{r}


plot_gpp_period <- ggplot(modis_period, aes(x = period, y = gpp_avg, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = gpp_avg - gpp_se, ymax = gpp_avg + gpp_se),
                width = 0.2, linewidth = 0.8) +

 
  scale_fill_manual(values = period_colors) + 

  
  labs(x = "Period", y = "", subtitle = "Period means ± SE" ) +#,        title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "none",
        #axis.text.x = element_text(angle = 45, vjust = 0.5),
        #axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = margin(2,2,2,2, "pt"))

plot_gpp_period 
```
### et

```{r}
et_normal <- modis_day %>% 
  filter(between(year, 2000, 2008)) %>% 
  summarize(etmean=mean(ET, na.rm=T)) %>% 
  pull(etmean)

et_anom <- modis_day %>% 
  filter(year>1999) %>% 
  mutate(season="annual", 
         pct_anomaly = 100 * (ET - et_normal) / et_normal,
         anomaly=ET-et_normal,
         cum_p_deficit = cumsum(anomaly)) #cumsum(pmax(0, -p_anom))

et_anom %>% 
  filter(year == 2019) %>%
  mutate( statement = sprintf("ET was %.1f%% %s than the 2000-2008 mean",
                             abs(pct_anomaly), ifelse(pct_anomaly < 0, "lower", "higher")))   
et_anom %>% 
  filter(between(year, 2009, 2020)) %>%
  summarize(mean_pct_anomaly = mean(pct_anomaly),
            mean_anomaly = mean(anomaly))  
```

```{r}



plot_et <-modisXseason %>% 
  bind_rows(modis_annual) %>% 
  filter(year>1999) %>% 
  ggplot( aes(x=year, y=et, fill=season)) +
  geom_col(position = position_dodge2(width = 0.8, padding = 0.05)) + 
  scale_x_continuous(breaks = seq(2000, 2025, 5), 
                     expand = expansion(add = c(0.1, 0.1))) +
  labs(x = "Year", y="ET (mm)", subtitle="")+
  scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +
  theme_classic() +
  theme(legend.position = "none",
        plot.margin = margin(2,2,2,2, "pt")) +

  geom_hline(yintercept = modis_normal$et_normal,
             linetype="dashed", color = "#5A5C8C") +
  geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5)
plot_et


```
```{r}


plot_et_period <- ggplot(modis_period, aes(x = period, y = et_avg, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = et_avg - et_se, ymax = et_avg + et_se),
                width = 0.2, linewidth = 0.8) +

  #scale_fill_manual(values = c("before" = "#7D593D", "during" = "lightgrey", "after" = "#AF9153")) +
  #scale_fill_viridis(discrete = TRUE, option = "cividis") +
  #paletteer::scale_fill_paletteer_d("rcartocolor::Safe") +
  scale_fill_manual(values = period_colors) + 
  
  labs(x = "Period", y = "") +#,       , subtitle = "Period means ± SE"  title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "none",
        #axis.text.x = element_text(angle = 45, vjust = 0.5),
        #axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = margin(2,2,2,2, "pt"))

plot_et_period 
```

### lai

have to take out ourliers
```{r}
range(modis_day$LAI, na.rm = T)


lai_annual <- modis_day %>% 
  filter(year!="",
         between(LAI, 0.01, 3.5)) %>% 
  group_by(year) %>% 
  summarize(
            lai=mean(LAI, na.rm=T))%>% 
  mutate(season="annual") 
range(lai_annual$lai, na.rm = T)

lai_normal <- lai_annual %>% 
  filter(between(year, 2002, 2008)) %>% 
  summarize(laimean=mean(lai, na.rm=T)) %>% 
  pull(laimean)


lai_anom <- lai_annual %>% 
  #filter(year>2001) %>% 
  mutate(season="annual", 
         pct_anomaly = 100 * (lai - lai_normal) / lai_normal,
         anomaly=lai-lai_normal,
         cum_p_deficit = cumsum(anomaly)) #cumsum(pmax(0, -p_anom))

lai_anom %>% 
  filter(year == 2019) %>%
  mutate( statement = sprintf("LAI was %.1f%% %s than the 2002-2008 mean",
                             abs(pct_anomaly), ifelse(pct_anomaly < 0, "lower", "higher")))   
lai_anom %>% 
  filter(between(year, 2009, 2020)) %>%
  summarize(mean_pct_anomaly = mean(pct_anomaly),
            mean_anomaly = mean(anomaly) ) 
```

```{r}
laiXseason <- modis_day %>% 
  filter(year!="",
         between(LAI, 0.01, 3.5)) %>% 
  mutate(season = if_else(month %in% breeding, "breeding", "nonbreeding")) %>% 
  group_by(year, season) %>% 
  summarize(lai=mean(LAI, na.rm=T)) %>% 
  ungroup() 

plot_lai <-laiXseason %>% 
  bind_rows(lai_annual) %>% 
  filter(year>1999) %>% 
  add_row(year = c(2000, 2000,2000, 2001, 2001, 2001), 
          season= c("annual", "breeding", "nonbreeding","annual", "breeding", "nonbreeding")) %>% 
  ggplot( aes(x=year, y=lai, fill=season)) +
  geom_col(position = position_dodge2(width = 0.8, padding = 0.05)) + 
  scale_y_continuous(limits = c(0, 2)) +
  scale_x_continuous(breaks = seq(2000, 2025, 5), 
                     expand = expansion(add = c(0.1, 0.1))) +
  labs(x = "Year", y=expression("LAI"~(m^2/m^2)), subtitle="")+
  scale_fill_manual(values=c("#5A5C8C", "#40B0A6", "#E1BE6A")) +
  theme_classic() +
  theme(legend.position = "none",plot.margin = margin(2,2,2,2, "pt")) +

  geom_hline(yintercept = lai_normal,
             linetype="dashed", color = "#5A5C8C") +
   geom_vline(xintercept = c(2008.5, 2019, 2020.5), 
             linetype = c("dotted", "solid", "dotted"), 
             color = c("black", "red", "black"), alpha = 0.5)
plot_lai


```


```{r}
lai_period <- lai_annual %>%  
  mutate(period = case_when(
      #year %in% period_clim ~ "longterm",
      year %in% period_before ~ "before",
      year %in% period_study  ~ "during",
      year %in% period_after  ~ "after",
      TRUE                   ~ NA_character_
    )) %>% 
  filter(year>2002) %>% 
  dplyr::group_by( period) %>%
  dplyr::summarize(n_years = n(),
                   lai_avg=mean(lai, na.rm = TRUE),
                   lai_sd=sd(lai, na.rm = TRUE),
                   lai_se = lai_sd / sqrt(n_years),
                   
                   .groups = "drop") %>% 
  add_row(period = "2019", 

          lai_avg = 0.2663043,
          n_years = 1)

lai_period$period<- factor(lai_period$period, 
                         levels=c( "before", "during", "after", "2019"),
                         labels=c("2000-2008", "2009-2020", "2021-2025", "2019"))

plot_lai_period <- ggplot(lai_period, aes(x = period, y = lai_avg, fill = period)) + #
  geom_col(alpha = 0.8, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = lai_avg - lai_se, ymax = lai_avg + lai_se),
                width = 0.2, linewidth = 0.8) +

  #scale_fill_manual(values = c("before" = "#7D593D", "during" = "lightgrey", "after" = "#AF9153")) +
  #scale_fill_viridis(discrete = TRUE, option = "cividis") +
  #paletteer::scale_fill_paletteer_d("rcartocolor::Safe") +
  scale_fill_manual(values = period_colors) + 
  
  labs(x = "Period", y = "" ) +#,        title = "Precipitation by Period") +
  theme_classic() +
  theme(legend.position = "none", 
        legend.direction = "vertical",
        legend.title = element_blank(),  # No "Period" title
        legend.text = element_text(size = 10),
        #axis.title.x = element_blank()
        axis.text.x = element_text(
               angle = 45,      # rotate 45 degrees
               hjust = 1,       # right-justify so labels line up with ticks
               vjust = 1        # adjust vertical position if needed
    ),
    plot.margin = margin(2,2,2,2, "pt"))

plot_lai_period 
```




## combine plots

Ecology: maximum allowable size (18 cm wide by 22 cm tall) 


```{r}
plots_modis <- ggarrange(plot_gpp + rremove("xlab"), 
                               plot_gpp_period + rremove("xlab"),
                               plot_et + rremove("xlab"),
                               plot_et_period + rremove("xlab"),
                               plot_lai, plot_lai_period,
                               nrow=3, ncol=2, align = "hv", widths = c(4,1),
                               common.legend = T, legend = "bottom") #
plots_modis
ggsave("./Revision/ecosystem_variation_v2.pdf", width=18, height=18, units = "cm", dpi=600)

plots_modis_legend <- ggarrange(plot_gpp + rremove("xlab"), 
                               plot_gpp_period + rremove("xlab"),
                               plot_et + rremove("xlab"),
                               plot_et_period + rremove("xlab"),
                               plot_lai, plot_lai_period,
                               nrow=3, ncol=2, align = "hv", widths = c(3,1),
                               legend = "bottom") #common.legend = T, 
ggsave("./Revision/ecosystem_variation_legend.pdf", width=18, height=15, units = "cm", dpi=600)
```


```{r plots modis}

plots_trend_modis <- ggarrange(plot_gpp + rremove("xlab"), 
           plot_et + rremove("xlab"), 
           plot_lai, 
           nrow=3, ncol=1, align = "hv", 
           common.legend = T, legend = "bottom")
plots_trend_modis

plots_period_modis <- ggarrange(plot_gpp_period + rremove("xlab"),
                          plot_et_period + rremove("xlab"),
                          plot_lai_period,
                          nrow=3, ncol=1, align = "hv", 
                          common.legend = T, legend = "bottom")
plots_period_modis      


ggarrange(plots_trend_modis, plots_period_modis, 
          ncol = 2, align = "h", 
          widths = c(2.5,1))


ggsave("./Revision/ecosystem_variation.pdf", width=18, height=15, units = "cm", dpi=600)


```





